---
title: "02.A Formatting growth curves from Synergy H1"
subtitle: "Quality control"
author: "Shane Hogle"
date: today
---

# Introduction

The ancestral (streptomycin sensitive) and evolved (streptomycin resistant) forms of HAMBI_1287 and HAMBI_1977 were grown on [Biolog Ecoplates](https://www.biolog.com/products/community-analysis-microplates/ecoplate/). These are basically a set of 32 different carbon compounds in 3 replicated sets on a 96 well plate. Microbial species or communities grown on these carbon compounds respire and reduce some dye to a purple color which is then measured on a Synergy H1 multimode microplate reader.

In this notebook we read the output from the Synergy H1 plate reader, join it to the samplesheets, smooth it, and inspect every well.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(fs)
library(ggforce)
library(slider)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw")
data_processed <- here::here("data", "processed", "02_biolog_ecoplates")

# make processed data directory if it doesn't exist
fs::dir_create(data_processed)
```

# Read and tidy Synergy output

Each plate is a separate acquisition on the Synergy H1. `read_synergy_xlsx()`
pulls the raw data sheet, sets the first reading as time zero, pivots to long
form, and zero-pads the well ids so they join to the samplesheets.

```{r}
plate01 <- read_synergy_xlsx(
  here::here(
    data_raw,
    "20250613_biolog_ecoplate_1287A",
    "1287A_raw_data.xlsx"
  )
) %>%
  # create a plate variable for later combining
  mutate(plate_name = "plate01")

plate02 <- read_synergy_xlsx(
  here::here(
    data_raw,
    "20250619_biolog_ecoplate_1287E_1977A",
    "1287E_1977A_rawdata.xlsx"
  )
) %>%
  mutate(plate_name = "plate02")

plate03 <- read_synergy_xlsx(
  here::here(
    data_raw,
    "20250626_biolog_ecoplate_1977A_1977E",
    "1977A_1977E_rawdata.xlsx"
  )
) %>%
  mutate(plate_name = "plate03")
```

# Combine, tidy, format

## Read sample metadata

```{r}
samplesheet01 <- readxl::read_xlsx(here::here(
  data_raw,
  "20250613_biolog_ecoplate_1287A",
  "samplesheet_1287A.xlsx"
)) %>%
  mutate(strainID = paste0("HAMBI_", strain)) %>%
  mutate(plate_name = "plate01")

samplesheet02 <- readxl::read_xlsx(here::here(
  data_raw,
  "20250619_biolog_ecoplate_1287E_1977A",
  "samplesheet_1287E_1977A.xlsx"
)) %>%
  mutate(strainID = paste0("HAMBI_", strain)) %>%
  mutate(plate_name = "plate02")

samplesheet03 <- readxl::read_xlsx(here::here(
  data_raw,
  "20250626_biolog_ecoplate_1977A_1977E",
  "samplesheet_1977A_1977E.xlsx"
)) %>%
  mutate(strainID = paste0("HAMBI_", strain)) %>%
  mutate(plate_name = "plate03")
```

## Join with metadata to remove unused samples

Grouping by plate and well, we calculate a five point centered rolling mean.
This reduces the jaggedness of the curves and also makes the resulting saved
data file smaller.

```{r}
ecoplate_gcurves_sm <- bind_rows(plate01, plate02, plate03) %>%
  left_join(
    bind_rows(samplesheet01, samplesheet02, samplesheet03),
    by = join_by(well, plate_name)
  ) %>%
  dplyr::group_by(plate_name, well) %>%
  dplyr::mutate(
    OD600_rollmean = slider::slide_dbl(OD600, mean, .before = 2, .after = 2)
  ) %>%
  ungroup() %>%
  relocate(OD600_rollmean, .after = "OD600")

# save output for later use
readr::write_tsv(
  ecoplate_gcurves_sm,
  here::here(data_processed, "gcurves_smooth.tsv")
)
```

# Inspect growth curves

```{r}
plotter <- function(id) {
  plotplate(
    ecoplate_gcurves_sm,
    dfxy = NULL,
    unsmoothed = TRUE,
    predicted = FALSE,
    plate = id,
    rows = 8,
    cols = 12,
    page = 1,
    scales = "fixed"
  )
}
```

## plate01

This plate contains three replicates for the ancestral form of HAMBI_1287.

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 12
#| fig-height: 9
#| fig-cap: Growth curves for the first ecoplate with the ancestral form of HAMBI_1287. X-axis is time in hours (48 hour incubation). Y axis is the absorbance scaled for each well. Blue line is smoothed with a moving average window of five points. Orange is non-smoothed.
plotter("plate01")
```

## plate02

This plate contains EVO_1287 (columns 1-8) and ANC_1977 (columns 9-12).

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 12
#| fig-height: 9
#| fig-cap: As in @fig-01 but for EVO_1287 (columns 1-8) and ANC_1977 (columns 9-12).
plotter("plate02")
```

## plate03

This plate contains ANC_1977 (columns 1-4) and EVO_1977 (columns 5-12).

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 12
#| fig-height: 9
#| fig-cap: As in @fig-01 but for ANC_1977 (columns 1-4) and EVO_1977 (columns 5-12).
plotter("plate03")
```

# Conclusions

Growth curves all look mostly reasonable. For just a couple of the carbon substrates there are some strange features (e.g., L-asparagine for HAMBI_1977). Can proceed with the analysis.
