---
title: "02.B Growth statistics on Biolog Ecoplates"
subtitle: "Spline fits and area under the curve"
author: "Shane Hogle"
date: today
---

# Introduction

Here we estimate maximum specific growth rate ($\mathrm{\mu_{max}}$) and total metabolic output (area under the growth curve) for every well of the Biolog Ecoplates formatted in the [prior step](01_format_growthcurves.qmd).

We use [growthrates](https://cran.r-project.org/web/packages/growthrates/index.html) to estimate $\mathrm{\mu_{max}}$. I have found this works a lot better than gcplyr and is more convenient than using another tool outside of R. Nonparametric estimation of growth rates by spline is very fast. Fitting to a model takes more time resources. Generally it is best to try multiple approaches and to visualize/check the data to make sure it makes sense ([Frontiers in Ecology and Evolution](https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2023.1313500/full)).

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(growthrates)
library(DescTools)
library(ggforce)
library(fs)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_interim <- here::here("data", "interim", "02_biolog_ecoplates")
data_processed <- here::here("data", "processed", "02_biolog_ecoplates")

# make data directories if they don't exist
fs::dir_create(data_interim)
fs::dir_create(data_processed)
```

## Read data

```{r}
gcurves_sm <- readr::read_tsv(
  here::here(data_processed, "gcurves_smooth.tsv")
) %>%
  # make uniq id
  mutate(id = paste0(plate_name, "|", well))
```

# Spline based estimates

Smoothing splines are a quick method to estimate maximum growth. The method is called nonparametric, because the growth rate is directly estimated from the smoothed data without being restricted to a specific model formula.

From the [growthrates documentation](https://cran.r-project.org/web/packages/growthrates/growthrates.pdf):

> The method was inspired by an algorithm of [Kahm et al. (2010)](https://www.jstatsoft.org/article/view/v033i07), with different settings and assumptions. In the moment, spline fitting is always done with log-transformed data, assuming exponential growth at the time point of the maximum of the first derivative of the spline fit. All the hard work is done by function `smooth.spline` from package stats, that is highly user configurable. Normally, smoothness is automatically determined via cross-validation. This works well in many cases, whereas manual adjustment is required otherwise, e.g. by setting `spar` to a fixed value \[0, 1\] that also disables cross-validation.

Fit splines to all growth curves.

```{r}
set.seed(45278)
many_spline <- growthrates::all_splines(
  OD600_rollmean ~ hours | id,
  data = gcurves_sm,
  spar = 0.5
)

readr::write_rds(many_spline, here::here(data_interim, "spline_fits.rds"))
```

Extract results.

```{r}
many_spline_res <- growthrates::results(many_spline)
```

Extract time of maximum specific growth rate for plotting.

```{r}
many_spline_xy <- purrr::map(many_spline@fits, \(x) {
  data.frame(x = x@xy[1], y = x@xy[2])
}) %>%
  purrr::list_rbind(names_to = "id") %>%
  separate_wider_delim(
    id,
    delim = "|",
    names = c("plate_name", "well")
  )
```

Extract slope of maximum specific growth rate for plotting.

```{r}
many_spline_fitted <- purrr::map(many_spline@fits, \(x) {
  data.frame(x@FUN(x@obs$time, x@par))
}) %>%
  purrr::list_rbind(names_to = "id") %>%
  dplyr::rename(hours = time, predicted = y) %>%
  dplyr::left_join(gcurves_sm, by = dplyr::join_by(id, hours)) %>%
  dplyr::group_by(id) %>%
  # this step makes sure we don't plot fits that go outside the range of the data
  dplyr::mutate(
    predicted = dplyr::if_else(
      dplyr::between(predicted, min(OD600_rollmean), max(OD600_rollmean)),
      predicted,
      NA_real_
    )
  ) %>%
  dplyr::ungroup()
```

# Area under growth curve (AUC)

Calculates AUC using the `DescTools` package.

```{r}
many_auc_res <- gcurves_sm %>%
  dplyr::summarize(
    auc = DescTools::AUC(hours, OD600_rollmean),
    max_od = max(OD600_rollmean),
    min_od = min(OD600_rollmean),
    .by = id
  ) %>%
  dplyr::left_join(
    dplyr::distinct(dplyr::select(gcurves_sm, strain:id)),
    by = join_by(id)
  ) %>%
  dplyr::select(-id)

# Write auc output
readr::write_tsv(
  many_auc_res,
  here::here(data_processed, "gcurve_auc_results.tsv")
)
```

Write spline results with strain info.

```{r}
many_spline_res %>%
  dplyr::left_join(
    dplyr::distinct(dplyr::select(gcurves_sm, strain:id)),
    by = join_by(id)
  ) %>%
  dplyr::select(-id) %>%
  readr::write_tsv(here::here(data_processed, "gcurve_spline_results.tsv"))
```

# Inspect the fits

```{r}
plotter <- function(plate) {
  plotplate(
    many_spline_fitted,
    dfxy = many_spline_xy,
    unsmoothed = FALSE,
    predicted = TRUE,
    plate = plate,
    rows = 8,
    cols = 12,
    page = 1,
    scales = "fixed"
  )
}
```

## plate01

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 12
#| fig-height: 8
#| fig-cap: Fitted growth curves for plate01 (ancestral HAMBI_1287), showing only smoothed data (blue line, five point moving average). Orange is slope of the maximum predicted growth rate from the first derivative of a smoothing spline. Red dot is hours and OD600 at which maximum growth rate is reached.
plotter("plate01")
```

## plate02

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for EVO_1287 (columns 1-8) and ANC_1977 (columns 9-12).
plotter("plate02")
```

## plate03

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for ANC_1977 (columns 1-4) and EVO_1977 (columns 5-12).
plotter("plate03")
```
