---
title: "02.C Strain growth on Biolog Ecoplates"
subtitle: "Carbon substrate use"
author: "Shane Hogle"
date: today
---

# Introduction

The [Biolog Ecoplates](https://www.biolog.com/products/community-analysis-microplates/ecoplate/) utilize tetrazolium reduction to measure the metabolic response of bacterial cells to different carbon substrates. Oxidized tetrazolium salt is water soluble, colorless in solution, and can freely diffuse into bacterial cells. Once inside the cell, tetrazolium is reduced by reducing agents like NADH an NADPH which are generated via cellular metabolism and catabolism. Once reduced tetrazolium is transformed into pink product. This assay measures the accumulation of reduced tetrazolium (colormetric absorbance at 590 nm) and hence the accumulation and concentration of reducing agents from cellular metabolism. This process will reflect both the concentration of cells in solution and per-capita concentration of reductant from central metabolism. There is a useful overview of the Tetrazolium colorometric procedure [available here.](https://fi.promega.com/resources/pubhub/is-your-mtt-assay-really-the-best-choice/)

In the [prior step](02_growth_curve_stats.qmd) we imported, smoothed, and calculated summary statistics for the species growth on Biolog Ecoplates. Here we will do some simple plots and analysis of the growth data on different carbon sources from the ecoplates.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(ggforce)
library(errors)
library(fs)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw")
data_processed <- here::here("data", "processed", "02_biolog_ecoplates")
figs <- here::here("output", "figures")

# make output directories if they don't exist
fs::dir_create(data_processed)
fs::dir_create(figs)
```

## Read data

Growth summary data (e.g., AUC and growth rates).

```{r}
many_auc_res <- readr::read_tsv(here::here(
  data_processed,
  "gcurve_auc_results.tsv"
))

many_spline_res <- readr::read_tsv(here::here(
  data_processed,
  "gcurve_spline_results.tsv"
))
```

Full growth curves.

```{r}
gcurves <- readr::read_tsv(here::here(data_processed, "gcurves_smooth.tsv")) %>%
  arrange(strainID, well, plate_name) %>%
  group_by(strainID, evolution, plate_name, `carbon source`, well) %>%
  mutate(id = dplyr::cur_group_id()) %>%
  group_by(strainID, evolution, `carbon source`) %>%
  mutate(
    replicate = case_when(
      id == min(id) ~ 1,
      id == max(id) ~ n_distinct(id),
      TRUE ~ 2
    )
  ) %>%
  mutate(replicate = LETTERS[replicate]) %>%
  dplyr::select(-id) %>%
  ungroup() %>%
  rename(carbon_source = `carbon source`) %>%
  mutate(
    hist = if_else(str_detect(evolution, "anc"), "Str Sens.", "Str Res.")
  )
```

Read which wells correspond to which compounds.

```{r}
cc <- readr::read_tsv(here::here(data_raw, "carbon_compound_map.tsv")) %>%
  dplyr::select(compound, class)
```

# Plot ecoplate growth curves

## HAMBI_1287

```{r}
p1287ecoplates_gcurves <- gcurves %>%
  filter(strainID == "HAMBI_1287") %>%
  ggplot() +
  ggplot2::geom_line(aes(
    x = hours,
    y = OD600_rollmean,
    color = hist,
    linetype = replicate,
    group = interaction(well, plate_name)
  )) +
  ggplot2::labs(x = "Hours", y = "OD600") +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 48, 12),
    labels = seq(0, 48, 12)
  ) +
  labs(
    y = "Metabolic performance (absorbance units, λ = 590 nm)",
    x = "Hours",
    color = "",
    lintype = "Replicate",
    title = "Citrobacter koseri HAMBI_1287"
  ) +
  facet_wrap(
    ~carbon_source,
    labeller = labeller(carbon_source = label_wrap_gen(20))
  ) +
  theme_mct()

ggsave(
  here::here(figs, "H1287_ecoplates_gcurves.svg"),
  plot = p1287ecoplates_gcurves,
  device = "svg",
  width = 7,
  height = 8
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 7
#| fig-height: 8
#| fig-cap: Full growth curves of *Citrobacter koseri* HAMBI_1287 on 31 different carbon substrates and water in Biolog Ecoplates over 48 hours of growth. Metabolic activity via tetrazolium reduction is colorometrically tracked with absorbance at 590 nm. Line type denotes the biological replicate, while line color denotes adaptation to the antibiotic streptomycin.
p1287ecoplates_gcurves
```

## HAMBI_1977

```{r}
p1977ecoplates_gcurves <- gcurves %>%
  filter(strainID == "HAMBI_1977") %>%
  ggplot() +
  ggplot2::geom_line(aes(
    x = hours,
    y = OD600_rollmean,
    color = hist,
    linetype = replicate,
    group = interaction(well, plate_name)
  )) +
  ggplot2::labs(x = "Hours", y = "OD600") +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 48, 12),
    labels = seq(0, 48, 12)
  ) +
  labs(
    y = "Metabolic performance (absorbance units, λ = 590 nm)",
    x = "Hours",
    color = "",
    lintype = "Replicate",
    title = "Pseudomonas chlororaphis HAMBI_1977"
  ) +
  facet_wrap(
    ~carbon_source,
    labeller = labeller(carbon_source = label_wrap_gen(20))
  ) +
  theme_mct()

ggsave(
  here::here(figs, "H1977_ecoplates_gcurves.svg"),
  plot = p1977ecoplates_gcurves,
  device = "svg",
  width = 7,
  height = 8
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 7
#| fig-height: 8
#| fig-cap: As in @fig-01 but for *Pseudomonas chlororaphis* HAMBI_1977.
p1977ecoplates_gcurves
```

# Plot ecoplate growth summaries

The procedure is:

1. for each replicate take the mean over all measurements within 0.05 units of the max
2. subtract water from all values
3. calculate mean and bootstrapped 95% CI across biological replicates

```{r}
gcurves_metabolic_pref <- gcurves %>%
  mutate(
    strainID2 = paste0(
      toupper(stringr::str_sub(evolution, start = 1, end = 1)),
      strain
    )
  ) %>%
  filter(hours > 24) %>%
  group_by(strainID2, carbon_source, replicate) %>%
  filter(OD600 >= max(OD600) - 0.05) %>%
  summarize(OD600 = mean(OD600)) %>%
  ungroup() %>%
  pivot_wider(names_from = carbon_source, values_from = "OD600") %>%
  pivot_longer(
    cols = c(-strainID2, -replicate, -water),
    names_to = "carbon_source",
    values_to = "OD600"
  ) %>%
  mutate(OD600_norm = if_else((OD600 - water) < 0, 0, OD600 - water)) %>%
  summarize(
    ggplot2::mean_cl_boot(OD600),
    .by = c(strainID2, carbon_source)
  )

gcurves_metabolic_pref_fmt <- gcurves_metabolic_pref %>%
  mutate(
    evolution = if_else(str_starts(strainID2, "A"), "Sens", "Res"),
    strainID2 = str_replace(strainID2, "A|E", "H")
  ) %>%
  pivot_wider(
    id_cols = c(strainID2, carbon_source),
    names_from = evolution,
    values_from = c(y, ymin, ymax)
  ) %>%
  mutate(
    y_Sens = errors::set_errors(y_Sens, ymax_Sens - y_Sens),
    y_Res = errors::set_errors(y_Res, ymax_Res - y_Res)
  ) %>%
  mutate(delta_sens_res = y_Sens - y_Res) %>%
  mutate(delta_sens_res_cl95 = errors::errors(delta_sens_res)) %>%
  mutate(delta_sens_res = errors::drop_errors(delta_sens_res)) %>%
  left_join(cc, by = join_by(carbon_source == compound)) %>%
  mutate(
    tot_met_output = errors::drop_errors(y_Sens + y_Res)
  )
```

Order the carbon substrates by compound class, then by the difference between
the sensitive and resistant forms of HAMBI_1287.

```{r}
carbon_levels <- gcurves_metabolic_pref_fmt %>%
  filter(strainID2 == "H1287") %>%
  arrange(delta_sens_res) %>%
  arrange(class) %>%
  distinct(carbon_source) %>%
  pull()
```

```{r}
pcarbonindv <- gcurves_metabolic_pref_fmt %>%
  mutate(carbon_source = factor(carbon_source, levels = carbon_levels)) %>%
  ggplot(
    aes(
      x = carbon_source,
      y = delta_sens_res,
      color = class
    ),
  ) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_point(aes(size = tot_met_output)) +
  geom_linerange(aes(
    ymin = delta_sens_res - delta_sens_res_cl95,
    ymax = delta_sens_res + delta_sens_res_cl95
  )) +
  labs(
    x = "",
    y = "Δ metabolic output (Sens - Res)",
    color = "Carbon class",
    size = "Total metabolic\noutput (Sens + Res)"
  ) +
  geom_line(aes(group = 1), alpha = 0.5) +
  coord_flip() +
  facet_grid(~strainID2)

ggsave(
  here::here(figs, "summary_metabolic_output.svg"),
  plot = pcarbonindv,
  device = "svg",
  width = 7,
  height = 7
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 7
#| fig-height: 7
#| fig-cap: Difference in metabolic performance between streptomycin sensitive and resistant forms of *Citrobacter koseri* 1287 and *Pseudomonas chlororaphis* 1977 when grown on 31 carbon substrates. Horizontal axis depicts the difference in maximum absorbance (AU, 590 nm) between the sensitive and resistant forms of each species over 48 hours. Point represents the difference in the mean over three biological replicates while the line range depicts the 95% confidence interval. Carbon sources are displayed on the vertical axis and organized by their parent molecular class (shown in color). Point size depicts the sum of the metabolic output of the sensitive and resistant forms.
pcarbonindv
```

# Appendix: superseded code

Code for plotting heatmaps of growth parameters, plotting maximum observed
absorbance, and clustering by similar responses. Retained for reference, not run.

```{r}
#| eval: false
# Some clustering to plot by more similar responses

# df2clust <- many_auc_res %>%
#   mutate(strainID2 = paste0(toupper(evolution), "_", strain)) %>%
#   summarize(y = min(max_od), .by = c(strainID2, `carbon source`)) %>%
#   pivot_wider(names_from = "carbon source", values_from = "y") %>%
#   pivot_longer(c(-strainID2, -water)) %>%
#   mutate(value_norm = value - water) %>%
#   pivot_wider(
#     id_cols = c(-water, -value),
#     names_from = "strainID2",
#     values_from = "value_norm"
#   ) %>%
#   as.data.frame() %>%
#   column_to_rownames(var = "name")

# # scale the dataframe for clustering
# df2clust_scaled <- scale(df2clust)

# # perform the hierarchcical clustering using euclidean distance and Ward's D
# hc <- hclust(dist(df2clust_scaled, method = "euclidean"), method = "ward.D2")

# # plot
# df2clust %>%
#   rownames_to_column(var = "carbon") %>%
#   pivot_longer(cols = -carbon) %>%
#   mutate(
#     carbon = factor(carbon, levels = hc$labels[hc$order]),
#     name = factor(
#       name,
#       levels = c("ANC_1287", "EVO_1287", "ANC_1977", "EVO_1977")
#     )
#   ) %>%
#   ggplot(aes(x = name, y = carbon)) +
#   geom_tile(aes(fill = value)) +
#   scale_x_discrete(guide = guide_axis(angle = 90)) +
#   scale_fill_viridis_c() +
#   scale_color_manual(values = c("white", "black"), guide = "none") +
#   labs(y = "Carbon substrate", fill = "Maximum\nabsorbance", x = "") +
#   coord_fixed() +
#   ggplot2::theme(
#     panel.grid = element_blank(),
#     panel.background = element_blank(),
#     strip.background = element_blank(),
#     panel.border = element_blank()
#   )

# many_auc_res %>%
#   mutate(strainID2 = paste0(toupper(evolution), "_", strain)) %>%
#   summarize(y = min(max_od), .by = c(strainID2, `carbon source`)) %>%
#   #summarize(ggplot2::mean_cl_boot(max_od), .by=c(strainID2, `carbon source`)) %>%
#   mutate(
#     strainID2 = factor(
#       strainID2,
#       levels = c("ANC_1287", "EVO_1287", "ANC_1977", "EVO_1977")
#     )
#   ) %>%
#   ggplot(aes(x = strainID2, y = `carbon source`)) +
#   geom_tile(aes(fill = y)) +
#   scale_x_discrete(guide = guide_axis(angle = 90)) +
#   scale_fill_viridis_c() +
#   scale_color_manual(values = c("white", "black"), guide = "none") +
#   labs(y = "Carbon substrate", fill = "Maximum\nabsorbance", x = "") +
#   coord_fixed() +
#   ggplot2::theme(
#     panel.grid = element_blank(),
#     panel.background = element_blank(),
#     strip.background = element_blank(),
#     panel.border = element_blank()
#   )

# # Plot AUC

# many_auc_res %>%
#   mutate(strainID2 = paste0(toupper(evolution), "_", strain)) %>%
#   summarize(
#     ggplot2::mean_cl_boot(auc),
#     .by = c(
#       strainID2,
#       `carbon source`
#     )
#   ) %>%
#   mutate(
#     strainID2 = factor(
#       strainID2,
#       levels = c("ANC_1287", "EVO_1287", "ANC_1977", "EVO_1977")
#     )
#   ) %>%
#   ggplot(aes(
#     x = strainID2,
#     y = `carbon source`
#   )) +
#   geom_tile(aes(fill = y)) +
#   scale_x_discrete(guide = guide_axis(angle = 90)) +
#   scale_fill_viridis_c() +
#   scale_color_manual(values = c("white", "black"), guide = "none") +
#   labs(y = "Carbon substrate", fill = "Area under the\ngrowth curve", x = "") +
#   coord_fixed() +
#   ggplot2::theme(
#     panel.grid = element_blank(),
#     panel.background = element_blank(),
#     strip.background = element_blank(),
#     panel.border = element_blank()
#   )

# # Plot growth rate
# many_spline_res %>%
#   mutate(strainID2 = paste0(toupper(evolution), "_", strain)) %>%
#   summarize(
#     ggplot2::mean_cl_boot(mumax),
#     .by = c(strainID2, `carbon source`)
#   ) %>%
#   mutate(
#     strainID2 = factor(
#       strainID2,
#       levels = c("ANC_1287", "EVO_1287", "ANC_1977", "EVO_1977")
#     )
#   ) %>%
#   ggplot(aes(x = strainID2, y = `carbon source`)) +
#   geom_tile(aes(fill = y)) +
#   scale_x_discrete(guide = guide_axis(angle = 90)) +
#   scale_fill_viridis_c() +
#   scale_color_manual(values = c("white", "black"), guide = "none") +
#   labs(y = "Carbon substrate", fill = "Growth rate\n(hr-1)", x = "") +
#   coord_fixed() +
#   ggplot2::theme(
#     panel.grid = element_blank(),
#     panel.background = element_blank(),
#     strip.background = element_blank(),
#     panel.border = element_blank()
#   )
```
