---
title: "05.C Growth curves on M9MM + single carbon"
subtitle: "Analysis and plots across streptomycin concentrations"
author: "Shane Hogle"
date: today
toc: true
toc-depth: 3
---

# Quick Jump:

### Citrobacter koseri 
- [amino acids](#sec-h1287-amino)
- [organic acids](#sec-h1287-organic)
- [sugars](#sec-h1287-sugar) 

### Pseudomonas chlororaphis
- [amino acids](#sec-h1977-amino)
- [organic acids](#sec-h1977-organic)
- [sugars](#sec-h1977-sugar)

# Introduction

Ancestral (streptomycin sensitive) and evolved (streptomycin resistant) forms of *Citrobacter koseri* HAMBI_1287 and *Pseudomonas chlororaphis* HAMBI_1977 were grown on 33 single carbon sources in M9 minimal medium, with and without streptomycin.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(growthrates)
library(fs)
library(glue)
library(marquee)
library(bayesboot)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_interim <- here::here("data", "interim", "05_one_carbon_gcurves")
data_processed <- here::here("data", "processed", "05_one_carbon_gcurves")
figs <- here::here("output", "figures")

# make output directories if they don't exist
fs::dir_create(data_interim)
fs::dir_create(data_processed)
fs::dir_create(figs)
```

# Load data

```{r}
growth_curves_sm_bl_filt <- read_tsv(here::here(
  data_processed,
  "growth_curves_sm_bl_filt_nr.tsv"
)) |>
  filter(strainID %in% c("H1287", "H1977")) |>
  filter(!is.na(OD600_rollmean))
```

# Processing

Take mean +- crude estimate of the 95% CI
```{r}
growth_curves_sm_bl_filt_95ci <- growth_curves_sm_bl_filt |>
  summarize(
    dirty_mean_ci(OD600_rollmean, ci = 0.95),
    .by = c(
      hours,
      strainID,
      evo_hist,
      streptomycin_ug_ml,
      carbon_source,
      carbon_class
    )
  )
```

# Plot curves

One panel of `carbon_source ~ streptomycin_ug_ml` facets per species per
`carbon_class`, so each plot stays small enough to read (carbon classes range
from 7 carbon sources for sugars up to 16 for amino acids). `theme_mct()` and
`histcols` (both from `R/generic.R`) keep the styling consistent with the rest
of the project.

```{r}
plot_gcurve_class <- function(data, species, class) {
  data |>
    filter(carbon_class == class & strainID == species) |>
    ggplot(aes(x = hours, y = mn)) +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi, fill = evo_hist), alpha = 0.5) +
    geom_line(aes(color = evo_hist)) +
    facet_grid(
      carbon_source ~ streptomycin_ug_ml,
      labeller = labeller(
        carbon_source = label_wrap_gen(15),
        streptomycin_ug_ml = \(x) paste0(x, " \u03bcg/ml")
      )
    ) +
    scale_color_manual(values = histcols) +
    scale_fill_manual(values = histcols) +
    labs(y = "OD_600", x = "Hours") +
    theme_mct() +
    theme(
      axis.text.y = marquee::element_marquee(size = 11),
      strip.text = element_text(size = 12),
      strip.text.y = element_text(angle = 0),
      legend.title = element_blank()
    )
}
```

## C. koseri HAMBI_1287 {#sec-h1287}

For coexistence, 1287 is the interesting case. Under drug, which strain is fitter flips with the carbon source: sugars, malate and amides favor EVO; other amino acids and succinate favor ANC.

### Amino acids {#sec-h1287-amino}

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-cap: "Mean OD600 (rolling average) over time for *C. koseri* HAMBI_1287 grown on 16 amino acid carbon sources, at 0 and 512 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 17
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1287", "amino_acid")
```

### Organic acids {#sec-h1287-organic}

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-cap: "Mean OD600 (rolling average) over time for *C. koseri* HAMBI_1287 grown on 11 organic acid carbon sources, at 0 and 512 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 12
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1287", "organic_acid")
```

### Sugars {#sec-h1287-sugar}

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-cap: "Mean OD600 (rolling average) over time for *C. koseri* HAMBI_1287 grown on 7 sugar carbon sources, at 0 and 512 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 8
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1287", "sugar")
```

## P. chlororaphis HAMBI_1977 {#sec-h1977}

### Amino acids {#sec-h1977-amino}

```{r}
#| echo: false
#| output: true
#| label: fig-04
#| fig-cap: "Mean OD600 (rolling average) over time for *P. chlororaphis* HAMBI_1977 grown on 16 amino acid carbon sources, at 0 and 10 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 17
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1977", "amino_acid")
```

### Organic acids {#sec-h1977-organic}

```{r}
#| echo: false
#| output: true
#| label: fig-05
#| fig-cap: "Mean OD600 (rolling average) over time for *P. chlororaphis* HAMBI_1977 grown on 11 organic acid carbon sources, at 0 and 10 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 12
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1977", "organic_acid")
```

### Sugars {#sec-h1977-sugar}

```{r}
#| echo: false
#| output: true
#| label: fig-06
#| fig-cap: "Mean OD600 (rolling average) over time for *P. chlororaphis* HAMBI_1977 grown on 7 sugar carbon sources, at 0 and 10 μg/ml streptomycin. Lines show the mean across replicate wells; ribbons show an estimate of the 95% CI (mean ± 1.96 SE). Orange = ancestral (streptomycin sensitive), purple = evolved (streptomycin resistant)."
#| fig-width: 7
#| fig-height: 8
plot_gcurve_class(growth_curves_sm_bl_filt_95ci, "H1977", "sugar")
```
