---
title: "05.A Carbon source platemap"
subtitle: "Parsing the platemap spreadsheets"
author: "Shane Hogle"
date: today
---

# Introduction

Converts the `Carbon_platemap*.xlsx` well maps in `data/raw/20260601_single_carbon_growth` into a single tidy platemap that [the next notebook](02_read_format_gcurves.qmd) joins onto the LogPhase 600 reads.

This code is ported from `scripts/10_one_carbon_gcurves_logphase/01_carbon_source_platemap_format.R` in [hambiEvoEnvCoexist](https://github.com/slhogle/hambiEvoEnvCoexist). The parsing is unchanged: the plates were run jointly, so the platemap describes all four species that were on them (HAMBI_0403, HAMBI_1287, HAMBI_1896, HAMBI_1977). Restricting to the two species this project cares about happens in the next notebook, so that the plate QC views still show every well as it was actually run.

## Parsing strategy

The spreadsheets are irregular: each sheet holds several plate blocks stacked vertically, and a block is only identifiable by scanning for its "Plate N" header. Within a block we read

* the well-column header row (1..12),
* the row-letter rows (A..H, read until column B goes empty),
* the "Medium/carbon source" row, and
* the "Streptomycin (ug/ml)" row.

The carbon source is resolved through the sheet's merged-cell ranges rather than forward-filled across the row. This matters: some blocks (uninoculated "Control" wells) deliberately have no medium entry and must not inherit the value from a neighbouring block.

## Cleaning

* "control" -> "Control" (inconsistent capitalisation in the sources).
* Strip the "M9MM with " prefix and " + Str" suffix from `carbon_source`; the streptomycin column already carries the "+Str" distinction.
* Attach the raw LogPhase filename for each plate via `plate_file_maps`.

## Per-file corrections

Three corrections cannot be derived from the spreadsheets and are applied explicitly in the "Per-file corrections" section below:

1. `Carbon_platemap_2`: source typo, "shicimic acid" -> "shikimic acid".
2. `Carbon_platemap_4`: the Streptomycin row is all zeros, including in the "+Str" well-columns, so columns 9-10 and 11-12 are set to 10 and 512.
3. `Carbon_platemap_5`: the well-column labels for the two right-hand blocks are transposed in the source sheet -- see `swap_column_blocks()`.

In every other platemap checked, the Streptomycin row is internally consistent (0 outside "+Str"/"Control" blocks, 10 for the ANC/EVO 1977 columns and 512 for the ANC/EVO 1287 columns inside a "+Str" block), so the spreadsheet values are trusted as-is.

::: {.callout-important}
## Correction 3 carries more weight here than upstream

In hambiEvoEnvCoexist only the streptomycin-free columns 1-4 were analysed, so the `Carbon_platemap_5` block transposition never touched a well that made it into a figure. **This project analyses the +Str wells**, which are exactly the columns that correction swaps. The evidence for the swap is in the reads themselves — columns 9-12 are empty across all 8 rows (never inoculated, i.e. the Control block) while columns 5-8 grew, and their per-column growth mirrors the strep-0 block in columns 1-4 strain for strain. Worth re-checking against the bench notebook before these results go into a manuscript.
:::

`Carbon_platemap_7.8` and `Carbon_platemap_7.8_2` differ structurally: each plate has no Streptomycin row at all (this batch doesn't test streptomycin dosing), and the uninoculated block's well cells are left empty rather than labelled "Control". Both are handled generically rather than as per-file corrections: a missing Streptomycin row parses as blank and later defaults to "0", and an empty well is flagged "blank" the same way a `strain == "Control"` well is.

# Setup

## Libraries

```{r}
library(tidyverse)
library(tidyxl)
library(openxlsx)
library(here)
library(fs)
```

## Cell access

```{r}
read_cell_matrices <- function(path) {
  cells <- tidyxl::xlsx_cells(path, sheets = 1)
  max_row <- max(cells$row)
  max_col <- max(cells$col)

  num_mat <- matrix(NA_real_, nrow = max_row, ncol = max_col)
  chr_mat <- matrix(NA_character_, nrow = max_row, ncol = max_col)
  idx <- as.matrix(cells[c("row", "col")])
  num_mat[idx] <- cells$numeric
  chr_mat[idx] <- cells$character

  list(num = num_mat, chr = chr_mat, max_row = max_row, max_col = max_col)
}

# Returns the numeric value if present, otherwise the character value. Scalar
# by necessity: the two matrices have different types.
cell_value <- function(mats, row, col) {
  out_of_range <- row < 1 || row > mats$max_row || col < 1 || col > mats$max_col
  if (out_of_range) {
    return(NA)
  }
  num_val <- mats$num[row, col]
  if (is.na(num_val)) mats$chr[row, col] else num_val
}
```

## Merged cell resolution

```{r}
read_merge_ranges <- function(path, mats) {
  wb <- openxlsx::loadWorkbook(path)
  refs <- stringr::str_match(wb$worksheets[[1]]$mergeCells, 'ref="([^"]+)"')[,
    2
  ]

  refs |>
    stringr::str_split(":") |>
    purrr::map(\(parts) {
      rows <- as.integer(stringr::str_extract(parts, "[0-9]+"))
      cols <- purrr::map_int(
        stringr::str_extract(parts, "[A-Z]+"),
        openxlsx::convertFromExcelRef
      )
      tibble::tibble(
        min_row = min(rows),
        max_row = max(rows),
        col = seq(min(cols), max(cols)),
        top_val = cell_value(mats, min(rows), min(cols))
      )
    }) |>
    purrr::list_rbind()
}

# For one sheet row, return the carbon source per well-column, preferring a
# merged-block value over the bare cell.
resolve_merged_row <- function(mats, merges, row, cols) {
  merged <- merges |>
    dplyr::filter(min_row <= row, row <= max_row, col %in% cols) |>
    dplyr::distinct(col, .keep_all = TRUE)

  tibble::tibble(col = cols) |>
    dplyr::mutate(
      bare_val = purrr::map_chr(
        col,
        \(cl) as.character(cell_value(mats, row, cl))
      )
    ) |>
    dplyr::left_join(dplyr::select(merged, col, top_val), by = "col") |>
    dplyr::mutate(
      carbon_source = dplyr::coalesce(as.character(top_val), bare_val)
    ) |>
    dplyr::select(col, carbon_source)
}
```

## Value cleaning

```{r}
clean_carbon_source <- function(x) {
  x <- stringr::str_trim(as.character(x))
  dplyr::case_when(
    is.na(x) ~ "",
    stringr::str_to_lower(x) == "control" ~ "Control",
    .default = x |>
      stringr::str_remove_all(stringr::fixed("M9MM with ")) |>
      stringr::str_remove_all(stringr::fixed(" + Str"))
  )
}

format_streptomycin <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  dplyr::case_when(
    is.na(x) ~ "",
    x == round(x) ~ sprintf("%.1f", x),
    .default = format(x, trim = TRUE, scientific = FALSE)
  )
}

# Strain labels look like "ANC1977" / "EVO1287"; anything else (e.g. "Control")
# is passed through as evo_hist with an empty strainID.
parse_evo_hist <- function(x) {
  dplyr::coalesce(
    stringr::str_extract(as.character(x), "^(ANC|EVO)"),
    as.character(x)
  )
}

parse_strain_id <- function(x) {
  dplyr::coalesce(
    stringr::str_match(as.character(x), "^(?:ANC|EVO)(\\d+)")[, 2],
    ""
  )
}
```

## Parsing

```{r}
parse_plate_block <- function(
  mats,
  merges,
  plate,
  medium_row,
  well_cols,
  row_rows,
  strep_row
) {
  carbon_by_col <- resolve_merged_row(
    mats,
    merges,
    medium_row,
    well_cols$col
  )

  tidyr::expand_grid(
    sheet_row = row_rows,
    col = well_cols$col
  ) |>
    dplyr::left_join(well_cols, by = "col") |>
    dplyr::left_join(carbon_by_col, by = "col") |>
    dplyr::mutate(
      plate = plate,
      row = purrr::map_chr(
        sheet_row,
        \(r) as.character(cell_value(mats, r, 2))
      ),
      strain = purrr::map2_chr(
        sheet_row,
        col,
        \(r, cl) as.character(cell_value(mats, r, cl))
      ),
      strep = purrr::map_chr(
        col,
        \(cl) as.character(cell_value(mats, strep_row, cl))
      ),
      well = stringr::str_c(row, well_column),
      evo_hist = parse_evo_hist(strain),
      strainID = parse_strain_id(strain),
      streptomycin_ug_ml = format_streptomycin(strep),
      carbon_source = clean_carbon_source(carbon_source)
    ) |>
    dplyr::select(
      plate,
      well,
      row,
      column = well_column,
      evo_hist,
      strainID,
      streptomycin_ug_ml,
      carbon_source
    )
}

# Locate the sheet rows that start each plate block.
find_plate_header_rows <- function(mats) {
  seq_len(mats$max_row) |>
    purrr::keep(\(r) {
      v <- cell_value(mats, r, 1)
      is.character(v) &&
        !is.na(v) &&
        stringr::str_starts(stringr::str_trim(v), "Plate")
    })
}

# The data rows of a block run from just below the header until column B is
# empty; the medium and streptomycin rows follow immediately after.
block_layout <- function(mats, header_row) {
  row_rows <- integer(0)
  rr <- header_row + 1
  while (rr <= mats$max_row && !is.na(cell_value(mats, rr, 2))) {
    row_rows <- c(row_rows, rr)
    rr <- rr + 1
  }
  list(row_rows = row_rows, medium_row = rr, strep_row = rr + 1)
}

parse_platemap <- function(xlsx_path) {
  mats <- read_cell_matrices(xlsx_path)
  merges <- read_merge_ranges(xlsx_path, mats)

  find_plate_header_rows(mats) |>
    purrr::map(\(header_row) {
      layout <- block_layout(mats, header_row)

      well_cols <- tibble::tibble(col = seq(3, mats$max_col)) |>
        dplyr::mutate(
          well_column = purrr::map_dbl(
            col,
            \(cl) suppressWarnings(as.numeric(cell_value(mats, header_row, cl)))
          )
        ) |>
        dplyr::filter(!is.na(well_column)) |>
        dplyr::mutate(well_column = as.integer(well_column))

      parse_plate_block(
        mats = mats,
        merges = merges,
        plate = stringr::str_trim(cell_value(mats, header_row, 1)),
        medium_row = layout$medium_row,
        well_cols = well_cols,
        row_rows = layout$row_rows,
        strep_row = layout$strep_row
      )
    }) |>
    purrr::list_rbind()
}
```

## Per-file correction helpers

```{r}
# Exchange two blocks of well-columns within a plate. Used where the source
# sheet's well-column header labels are transposed relative to the physical
# plate: the labels parsed under `from` actually describe the wells in `to` and
# vice versa, so the records are relabelled symmetrically.
#
# `well` is rebuilt from `row` and the corrected column so that it stays
# consistent with `column`.
swap_column_blocks <- function(records, from, to, plates = NULL) {
  stopifnot(length(from) == length(to))

  in_scope <- if (is.null(plates)) TRUE else records$plate %in% plates

  records |>
    dplyr::mutate(
      column = dplyr::case_when(
        !in_scope ~ column,
        column %in% from ~ to[match(column, from)],
        column %in% to ~ from[match(column, to)],
        .default = column
      ),
      well = stringr::str_c(row, column)
    )
}
```

## Read and correct the platemaps

```{r}
data_dir <- here::here("data", "raw", "20260601_single_carbon_growth")

source_records <- list(
  carbonsources_12 = "12_carbonsources.xlsx",
  hydroxybenzoic_salicylic = "4-hydrobenzoic acid_salicylic acid.xlsx",
  platemap_1 = "Carbon_platemap_1.xlsx",
  platemap_2 = "Carbon_platemap_2.xlsx",
  platemap_3 = "Carbon_platemap_3.xlsx",
  platemap_4 = "Carbon_platemap_4.xlsx",
  platemap_5 = "Carbon_platemap_5.xlsx",
  platemap_7_8 = "Carbon_platemap_7.8.xlsx",
  platemap_7_8_2 = "Carbon_platemap_7.8_2.xlsx",
  starch_malate = "Starch_malate.xlsx"
) |>
  purrr::map(\(f) parse_platemap(file.path(data_dir, f)))
```

## Per-file corrections

```{r}
# Not reproducible from the spreadsheets themselves (see header note).

# (1) Carbon_platemap_2: source spreadsheet typo
source_records$platemap_2 <- source_records$platemap_2 |>
  dplyr::mutate(
    carbon_source = stringr::str_replace(
      carbon_source,
      "^shicimic acid$",
      "shikimic acid"
    )
  )

# (2) Carbon_platemap_4: Streptomycin row is all zeros, including in the "+Str"
# well-columns (9-10 = ANC/EVO 1977, 11-12 = ANC/EVO 1287)
source_records$platemap_4 <- source_records$platemap_4 |>
  dplyr::mutate(
    streptomycin_ug_ml = dplyr::case_when(
      column %in% c(9, 10) ~ "10.0",
      column %in% c(11, 12) ~ "512.0",
      .default = streptomycin_ug_ml
    )
  )

# (3) Carbon_platemap_5: the well-column labels of the "+Str" block and the
# uninoculated "Control" block are transposed in the source sheet. As parsed,
# columns 5-8 come out as Control and columns 9-12 as the "+Str" strain wells.
# The reads show the opposite, unambiguously:
#
#   * columns 9-12 are empty across all 8 rows -- never inoculated, i.e. the
#     Control block;
#   * columns 5-8 grew, and their per-column growth mirrors the strep-0 block
#     in columns 1-4 strain for strain (1977 ANC/EVO, then 1287 ANC/EVO), which
#     is exactly the "+Str" block layout.
#
# So the two blocks are exchanged: 5<->9, 6<->10, 7<->11, 8<->12. Columns 1-4
# (the streptomycin-free wells this project actually analyses) are untouched.
source_records$platemap_5 <- source_records$platemap_5 |>
  swap_column_blocks(from = 5:8, to = 9:12)
```

## Output

```{r}
plate_file_maps <- list(
  carbonsources_12 = c(
    "1" = "Carbon_sources_5_1_06-heinä-2026 09-01-04.txt",
    "2" = "Carbon_sources_5_2_06-heinä-2026 09-01-04.txt",
    "3" = "Carbon_sources_5_3_06-heinä-2026 09-01-04.txt",
    "4" = "Carbon_sources_5_4_06-heinä-2026 09-01-04.txt"
  ),
  hydroxybenzoic_salicylic = c(
    "1" = "Carbon_sources_9_2_09-heinä-2026 14-42-47.txt",
    "2" = "Carbon_sources_9_3_09-heinä-2026 14-42-47.txt"
  ),
  platemap_1 = c(
    "1" = "Carbon_sources_1_1_08-kesä-2026 12-51-22.txt",
    "2" = "Carbon_sources_1_2_08-kesä-2026 12-51-22.txt",
    "3" = "Carbon_sources_1_3_08-kesä-2026 12-51-22.txt",
    "4" = "Carbon_sources_1_4_08-kesä-2026 12-51-22.txt"
  ),
  platemap_2 = c(
    "1" = "Carbon_sources_2_1_22-kesä-2026 08-41-34.txt",
    "2" = "Carbon_sources_2_2_22-kesä-2026 08-41-34.txt",
    "3" = "Carbon_sources_2_3_22-kesä-2026 08-41-34.txt",
    "4" = "Carbon_sources_2_4_22-kesä-2026 08-41-34.txt"
  ),
  platemap_3 = c(
    "1" = "Carbon_sources_3_1_26-kesä-2026 12-49-10.txt",
    "2" = "Carbon_sources_3_2_26-kesä-2026 12-49-10.txt",
    "3" = "Carbon_sources_3_3_26-kesä-2026 12-49-10.txt",
    "4" = "Carbon_sources_3_4_26-kesä-2026 12-49-10.txt"
  ),
  platemap_4 = c(
    "1" = "Carbon_sources_4_1_30-kesä-2026 10-07-18.txt",
    "2" = "Carbon_sources_4_2_30-kesä-2026 10-07-18.txt"
  ),
  platemap_5 = c(
    "1" = "Carbon_sources_9_1_09-heinä-2026 14-42-47.txt"
  ),
  platemap_7_8 = c(
    "1" = "Carbon_sources_10_1_10-elo-2026 09-09-08.txt",
    "2" = "Carbon_sources_10_2_10-elo-2026 09-09-08.txt"
  ),
  platemap_7_8_2 = c(
    "3" = "Carbon_sources_10_3_10-elo-2026 09-09-08.txt",
    "4" = "Carbon_sources_10_4_10-elo-2026 09-09-08.txt"
  ),
  starch_malate = c(
    "1" = "Carbon_sources_7_1_06-heinä-2026 09-01-31.txt"
  )
)

carbon_source_classes <- tibble::tribble(
  ~carbon_source          , ~class         ,
  "4-Hydroxybenzoic acid" , "organic_acid" ,
  "glucose"               , "sugar"        ,
  "sucrose"               , "sugar"        ,
  "salicylic acid"        , "organic_acid" ,
  "shikimic acid"         , "organic_acid" ,
  "L-cysteine"            , "amino_acid"   ,
  "L-valine"              , "amino_acid"   ,
  "taurine"               , "amino_acid"   ,
  "histidine"             , "amino_acid"   ,
  "leucine"               , "amino_acid"   ,
  "phenylalanine"         , "amino_acid"   ,
  "xylose"                , "sugar"        ,
  "starch"                , "sugar"        ,
  "cellobiose"            , "sugar"        ,
  "succinic acid"         , "organic_acid" ,
  "malate"                , "organic_acid" ,
  "asparagine"            , "amino_acid"   ,
  "D-mannitol"            , "sugar"        ,
  "lactic acid"           , "organic_acid" ,
  "sodium acetate"        , "organic_acid"
)

attach_plate_file <- function(records, plate_file_map) {
  records |>
    dplyr::mutate(
      plate_file = plate_file_map[stringr::str_extract(plate, "[0-9]+")]
    ) |>
    dplyr::select(
      plate,
      well,
      row,
      column,
      evo_hist,
      strainID,
      streptomycin_ug_ml,
      carbon_source,
      plate_file
    )
}

master_tidy <- source_records |>
  purrr::imap(
    \(records, nm) attach_plate_file(records, plate_file_maps[[nm]])
  ) |>
  purrr::list_rbind() |>
  dplyr::mutate(
    # well format must match the samplesheet, e.g. "A01" not "A1"
    well = stringr::str_c(
      stringr::str_extract(well, "^[A-H]"),
      stringr::str_pad(
        stringr::str_extract(well, "\\d+"),
        width = 2,
        pad = "0",
        side = "left"
      )
    ),
    strainID = dplyr::if_else(
      strainID != "",
      stringr::str_c(
        "H",
        stringr::str_pad(strainID, width = 4, pad = "0", side = "left")
      ),
      strainID
    ),
    streptomycin_ug_ml = stringr::str_remove(streptomycin_ug_ml, "\\.0")
  ) |>
  # uninoculated wells are flagged "blank" across every descriptive column.
  # Most sheets label these wells "Control" in the strain cell;
  # Carbon_platemap_7.8 and _2 instead leave the strain cell empty, hence the
  # is.na() check.
  dplyr::mutate(
    dplyr::across(
      c(evo_hist, strainID, streptomycin_ug_ml, carbon_source),
      \(x) {
        dplyr::if_else(
          is.na(evo_hist) | evo_hist == "Control" | carbon_source == "Control",
          "blank",
          x
        )
      }
    )
  ) |>
  dplyr::mutate(
    streptomycin_ug_ml = dplyr::if_else(
      streptomycin_ug_ml == "",
      "0",
      streptomycin_ug_ml
    ),
    # fixing inconsistent carbon source names across sheets
    carbon_source = dplyr::case_when(
      carbon_source == "valine" ~ "L-valine",
      carbon_source == "malic acid (malate)" ~ "malate",
      carbon_source == "4-hydrobenzoic acid" ~ "4-Hydroxybenzoic acid",
      carbon_source == "cysteine" ~ "L-cysteine",
      carbon_source == "arginine" ~ "L-arginine",
      .default = carbon_source
    )
  ) |>
  dplyr::left_join(carbon_source_classes, by = dplyr::join_by(carbon_source))

# Written to data/interim rather than alongside the source spreadsheets: this is
# a derived file, and nothing under data/raw should be modified.
data_interim <- here::here("data", "interim", "05_one_carbon_gcurves")
fs::dir_create(data_interim)

readr::write_tsv(
  master_tidy,
  here::here(data_interim, "platemap_master_tidy.tsv"),
  na = "",
  eol = "\n"
)
```
