---
title: "05.B Growth curves on M9MM + single carbon"
subtitle: "Quality control"
author: "Shane Hogle"
date: today
---

# Introduction

Each strain was grown in M9 minimal medium with a single compound supplied as the sole carbon source, at three streptomycin concentrations, and read on a Biotek Logphase 600. This notebook joins the LogPhase reads to the platemap from the [previous step](01_carbon_source_platemap_format.qmd), smooths them, and inspects every plate.

The plates were run jointly with [hambiEvoEnvCoexist](https://github.com/slhogle/hambiEvoEnvCoexist) and carry four species. The QC views below show every well as it was run; the saved table is restricted to the two species this project uses (HAMBI_1287 and HAMBI_1977) plus the uninoculated blanks.

Streptomycin dosing is per species — HAMBI_1977 wells were dosed at 10 μg/ml and HAMBI_1287 wells at 512 μg/ml, alongside the streptomycin-free wells. So each species has two levels, not a shared gradient.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(fs)
library(ggforce)
library(slider)
library(stringi)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw", "20260601_single_carbon_growth")
data_interim <- here::here("data", "interim", "05_one_carbon_gcurves")
data_processed <- here::here("data", "processed", "05_one_carbon_gcurves")

# make output directories if they don't exist
fs::dir_create(data_interim)
fs::dir_create(data_processed)
```

# Read

```{r}
# read in the metadata plate map from the previous step
platemap <- read_tsv(here::here(data_interim, "platemap_master_tidy.tsv")) |>
  # filenames on disk use NFD Unicode normalization (e.g. "ä" stored as
  # decomposed a + combining diaeresis, U+0308) while this table stores
  # the precomposed NFC form; normalize to NFD here so paths built
  # downstream match the literal bytes on disk (Linux filesystems don't
  # auto-normalize, unlike macOS/HFS+)
  mutate(plate_file = stri_trans_nfd(plate_file))
```

```{r}
# read in OD data
ods_combined <- platemap |>
  select(plate_file) |>
  distinct() |>
  mutate(file1 = here::here(data_raw, paste(plate_file))) |>
  pull(file1) |>
  map(\(x) read_logphase_txt(x, skip = 28)) |>
  list_rbind()
```

# Format

```{r}
# join combined optical density data with platemap
ods_combined_md <- left_join(ods_combined, platemap) |>
  # add a unique id
  group_by(plate_file, plate) |>
  mutate(plate_name = cur_group_id()) |>
  ungroup()
```

```{r}
# smooth growth curves
ods_combined_md_sm <- ods_combined_md |>
  dplyr::group_by(plate_name, well) |>
  arrange(plate_file, plate, well) |>
  # subtract od at first time point from gcurve so it starts at 0 OD
  mutate(OD600 = OD600 - first(OD600)) |>
  # set negative values to 0
  mutate(OD600 = if_else(OD600 < 0, 0, OD600)) |>
  dplyr::mutate(
    OD600_rollmean = slider::slide_dbl(OD600, mean, .before = 2, .after = 2)
  ) |>
  ungroup() |>
  relocate(OD600_rollmean, .after = "OD600")
```

# Plot

Inspecting individual plates. There are 26 plates.

::: {.panel-tabset}

## Plate 1

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 1. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 1,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 1)
```

## Plate 2

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 2. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 2,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 2)
```

## Plate 3

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 3. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 3,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 3)
```

## Plate 4

```{r}
#| echo: false
#| output: true
#| label: fig-04
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 4. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 4,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 4)
```

## Plate 5

```{r}
#| echo: false
#| output: true
#| label: fig-05
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 5. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 5,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 5)
```

## Plate 6

```{r}
#| echo: false
#| output: true
#| label: fig-06
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 6. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 6,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 6)
```

## Plate 7

```{r}
#| echo: false
#| output: true
#| label: fig-07
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 7. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 7,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 7)
```

## Plate 8

```{r}
#| echo: false
#| output: true
#| label: fig-08
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 8. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 8,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 8)
```

## Plate 9

```{r}
#| echo: false
#| output: true
#| label: fig-09
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 9. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 9,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 9)
```

## Plate 10

```{r}
#| echo: false
#| output: true
#| label: fig-10
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 10. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 10,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 10)
```

## Plate 11

```{r}
#| echo: false
#| output: true
#| label: fig-11
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 11. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 11,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 11)
```

## Plate 12

```{r}
#| echo: false
#| output: true
#| label: fig-12
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 12. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 12,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 12)
```

## Plate 13

```{r}
#| echo: false
#| output: true
#| label: fig-13
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 13. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 13,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 13)
```

## Plate 14

```{r}
#| echo: false
#| output: true
#| label: fig-14
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 14. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 14,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 14)
```

## Plate 15

```{r}
#| echo: false
#| output: true
#| label: fig-15
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 15. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 15,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 15)
```

## Plate 16

```{r}
#| echo: false
#| output: true
#| label: fig-16
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 16. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 16,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 16)
```

## Plate 17

```{r}
#| echo: false
#| output: true
#| label: fig-17
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 17. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 17,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 17)
```

## Plate 18

```{r}
#| echo: false
#| output: true
#| label: fig-18
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 18. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 18,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 18)
```

## Plate 19

```{r}
#| echo: false
#| output: true
#| label: fig-19
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 19. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 19,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 19)
```

## Plate 20

```{r}
#| echo: false
#| output: true
#| label: fig-20
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 20. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 20,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 20)
```

## Plate 21

```{r}
#| echo: false
#| output: true
#| label: fig-21
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 21. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 21,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 21)
```

## Plate 22

```{r}
#| echo: false
#| output: true
#| label: fig-22
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 22. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 22,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 22)
```

## Plate 23

```{r}
#| echo: false
#| output: true
#| label: fig-23
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 23. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 23,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 23)
```

## Plate 24

```{r}
#| echo: false
#| output: true
#| label: fig-24
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 24. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 24,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 24)
```

## Plate 25

```{r}
#| echo: false
#| output: true
#| label: fig-25
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 25. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 25,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 25)
```

## Plate 26

```{r}
#| echo: false
#| output: true
#| label: fig-26
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Logphase 600 plate_name = 26. Blue line is smoothed with a moving average window of five points. Orange dashed is non-smoothed."
plotplate(
  ods_combined_md_sm,
  dfxy = NULL,
  unsmoothed = TRUE,
  predicted = FALSE,
  plate = 26,
  rows = 8,
  cols = 12,
  page = 1,
  scales = "fixed"
)
```

```{r}
#| echo: false
#| output: true
plate_summary(ods_combined_md_sm, 26)
```

:::

# Save

Remove these wells after inspecting the plots above — they look suspect.

`plate_name` is a `cur_group_id()` over `(plate_file, plate)`, so its integers are
reassigned whenever a new `plate_file` sorts ahead of existing ones alphabetically
(e.g. "Carbon_sources_10_..." sorts before "Carbon_sources_1_..."). Keying this list
on `plate_name` would silently blacklist the wrong wells the next time a batch is
added, so it's keyed on `(plate_file, plate, well)` instead, which is stable.

```{r}
blacklist <- tribble(
  ~plate_file                                     , ~plate    , ~well ,
  "Carbon_sources_1_3_08-kesä-2026 12-51-22.txt"  , "Plate 3" , "F01" ,
  "Carbon_sources_3_4_26-kesä-2026 12-49-10.txt"  , "Plate 4" , "H01" ,
  "Carbon_sources_3_4_26-kesä-2026 12-49-10.txt"  , "Plate 4" , "H02" ,
  "Carbon_sources_9_3_09-heinä-2026 14-42-47.txt" , "Plate 2" , "A01" ,
  "Carbon_sources_9_3_09-heinä-2026 14-42-47.txt" , "Plate 2" , "G02" ,
  "Carbon_sources_9_3_09-heinä-2026 14-42-47.txt" , "Plate 2" , "E04"
) |>
  # match the NFD normalization applied to plate_file above
  mutate(plate_file = stri_trans_nfd(plate_file))
```

Restrict to the species used in this project. HAMBI_0403 and HAMBI_1896 shared
the plates but belong to the sister project; the uninoculated blanks are kept so
that downstream steps can still see them.

```{r}
anti_join(ods_combined_md_sm, blacklist) |>
  filter(strainID %in% c("H1287", "H1977", "blank")) |>
  write_tsv(here::here(data_processed, "one_carbon_gcurves_logphase.tsv"))
```
