---
title: "05.C Growth curves on M9MM + single carbon"
subtitle: "Analysis and plots across streptomycin concentrations"
author: "Shane Hogle"
date: today
---

# Introduction

Ancestral (streptomycin sensitive) and evolved (streptomycin resistant) forms of *Citrobacter koseri* HAMBI_1287 and *Pseudomonas chlororaphis* HAMBI_1977 were grown on 20 single carbon sources in M9 minimal medium, with and without streptomycin.

This analysis is ported from `scripts/10_one_carbon_gcurves_logphase/03_analyze_gcurves.qmd` in [hambiEvoEnvCoexist](https://github.com/slhogle/hambiEvoEnvCoexist), which restricted itself to the streptomycin-free wells. Here we keep every streptomycin concentration.

::: {.callout-note}
## The streptomycin design is nested inside species, not crossed with it

Dosing is per species: HAMBI_1977 wells were dosed at 10 μg/ml and HAMBI_1287 wells at 512 μg/ml. There is no concentration at which both species were tested, so the streptomycin panels below compare ANC against EVO *within* a species, never one species against the other at matched dose.

Four carbon sources — D-mannitol, L-arginine, lactic acid and sodium acetate — were only ever run without streptomycin. They appear as gaps in the +Str panels rather than being dropped, so that rows stay aligned across panels.
:::

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(growthrates)
library(fs)
library(glue)
library(marquee)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_interim <- here::here("data", "interim", "05_one_carbon_gcurves")
data_processed <- here::here("data", "processed", "05_one_carbon_gcurves")
figs <- here::here("output", "figures")

# make output directories if they don't exist
fs::dir_create(data_interim)
fs::dir_create(data_processed)
fs::dir_create(figs)
```

# Load data

```{r}
df <- read_tsv(here::here(
  data_processed,
  "one_carbon_gcurves_logphase.tsv"
)) |>
  mutate(plate_name = as.character(plate_name))
```

# Processing

## Growth rates per well (spline based estimate)

Using the tool [`growthrates`](https://cran.r-project.org/web/packages/growthrates/index.html) to estimate mu_max. I have found this works a lot better than gcplyr and is more convenient than using another tool outside of R. Nonparametric estimate growth rates by spline is very fast. Fitting to a model takes more time resources. [Generally it is best to try multiple approaches and to visualize/check the data to make sure it makes sense.](https://www.frontiersin.org/journals/ecology-and-evolution/articles/10.3389/fevo.2023.1313500/full)

Unlike upstream, the streptomycin-dosed wells are kept here.

```{r}
gcurves_thin_sm <- df |>
  # exclude blanks, but keep every streptomycin concentration
  filter(strainID != "blank") |>
  # need to add a small floor for numerical stability for the spline fit
  mutate(
    OD600_rollmean_spline = if_else(OD600_rollmean == 0, 1e-3, OD600_rollmean)
  ) |>
  # make uniq id
  mutate(id = paste0(plate_name, "|", well))
```

Smoothing splines are a quick method to estimate maximum growth. The method is
called nonparametric, because the growth rate is directly estimated from the
smoothed data without being restricted to a specific model formula. From
[growthrates documentation:](https://cran.r-project.org/web/packages/growthrates/growthrates.pdf)

> The method was inspired by an algorithm of [Kahm et al.
> (2010)](https://www.jstatsoft.org/article/view/v033i07), with different
> settings and assumptions. In the moment, spline fitting is always done with
> log-transformed data, assuming exponential growth at the time point of the
> maximum of the first derivative of the spline fit. All the hard work is done
> by function smooth.spline from package stats, that is highly user
> configurable. Normally, smoothness is automatically determined via
> cross-validation. This works well in many cases, whereas manual adjustment
> is required otherwise, e.g. by setting spar to a fixed value \[0, 1\] that
> also disables cross-validation.

```{r}
set.seed(45278)
many_spline <- growthrates::all_splines(
  OD600_rollmean_spline ~ hours | id,
  data = gcurves_thin_sm,
  spar = 0.5
)

readr::write_rds(
  many_spline,
  here::here(data_interim, "one_carbon_spline_fits.rds")
)
```

```{r}
# get results
gr_per_well <- growthrates::results(many_spline) |>
  tidyr::separate('id', into = c("plate_name", "well")) |>
  left_join(distinct(dplyr::select(
    df,
    -seconds,
    -hours,
    -OD600,
    -OD600_rollmean
  )))
```

## Compute K per well

Simply take the maximum observed OD600 value to get a "dumb" empirical estimate of the carrying capacity.

```{r}
k_per_well <- df |>
  filter(strainID != "blank") |>
  arrange(plate_file, well, hours) |>
  summarize(
    max_od = max(OD600_rollmean),
    .by = c(
      plate_name,
      well,
      strainID,
      evo_hist,
      streptomycin_ug_ml,
      carbon_source,
      class
    )
  )
```

## Compute AUC per well

Area under the growth curve (AUC) is computed per unique well using the
trapezoidal rule over time in hours. `OD600_rollmean` values are used to reduce noise.

```{r}
auc_per_well <- df |>
  filter(strainID != "blank") |>
  arrange(plate_file, well, hours) |>
  summarize(
    auc = trap_auc(hours, OD600_rollmean),
    .by = c(
      plate_name,
      well,
      strainID,
      evo_hist,
      streptomycin_ug_ml,
      carbon_source,
      class
    )
  )
```

## Write growth summaries to disc

`streptomycin_ug_ml` is carried into the summary table here — upstream it was a
constant 0 and could be dropped.

```{r}
myformatter <- function(df, grvar) {
  df |>
    dplyr::select(
      species = strainID,
      evo_hist,
      streptomycin_ug_ml,
      carbon_source,
      carbon_class = class,
      {{ grvar }}
    ) |>
    mutate(
      genotype = str_sub(evo_hist, 1, 1),
    ) |>
    arrange(
      species,
      genotype,
      streptomycin_ug_ml,
      carbon_source,
      carbon_class
    ) |>
    group_by(
      species,
      genotype,
      streptomycin_ug_ml,
      carbon_source,
      carbon_class
    ) |>
    mutate(replicate = LETTERS[1:n()]) |>
    ungroup() |>
    dplyr::select(
      species,
      genotype,
      streptomycin_ug_ml,
      carbon_source,
      carbon_class,
      replicate,
      {{ grvar }}
    )
}

left_join(myformatter(gr_per_well, mumax), myformatter(k_per_well, max_od)) |>
  left_join(myformatter(auc_per_well, auc)) |>
  rename(
    growth_rate = mumax,
    carrying_capacity = max_od,
    area_under_growth_curve = auc
  ) |>
  write_tsv(here::here(data_processed, "growth_summaries_carbon.tsv"))
```

## Bootstrap summary by strainID, evo_hist, streptomycin_ug_ml, carbon_source

```{r}
#| output: false
#| warning: false
set.seed(7391)
N_BOOT <- 2000

auc_summary <- auc_per_well |>
  summarize(
    boot_mean_ci(auc),
    .by = c(strainID, evo_hist, streptomycin_ug_ml, carbon_source, class)
  )

k_summary <- k_per_well |>
  summarize(
    boot_mean_ci(max_od),
    .by = c(strainID, evo_hist, streptomycin_ug_ml, carbon_source, class)
  )

gr_summary <- gr_per_well |>
  summarize(
    boot_mean_ci(mumax),
    .by = c(strainID, evo_hist, streptomycin_ug_ml, carbon_source, class)
  )
```

Replicate counts per point, so the figure captions do not have to assert one.

```{r}
#| echo: false
#| output: true
auc_per_well |>
  count(strainID, streptomycin_ug_ml, name = "wells") |>
  left_join(
    auc_per_well |>
      count(strainID, streptomycin_ug_ml, evo_hist, carbon_source) |>
      summarize(
        reps_per_point = paste(sort(unique(n)), collapse = ", "),
        .by = c(strainID, streptomycin_ug_ml)
      ),
    by = join_by(strainID, streptomycin_ug_ml)
  )
```

# Plot results

```{r}
# order of the carbon sources
carbon_levels <- rev(c(
  "glucose",
  "xylose",
  "sucrose",
  "cellobiose",
  "starch",
  "D-mannitol",
  "succinic acid",
  "malate",
  "4-Hydroxybenzoic acid",
  "shikimic acid",
  "salicylic acid",
  "lactic acid",
  "sodium acetate",
  "histidine",
  "leucine",
  "L-valine",
  "L-cysteine",
  "phenylalanine",
  "taurine",
  "L-arginine"
))
```

Upstream the axis-label colours were a hard-coded run of 7/7/6. Here they are
looked up from the `class` column instead, so the colouring cannot silently
drift out of step with `carbon_levels`.

```{r}
class_colors <- c(
  "sugar" = "#17A769",
  "organic_acid" = "#E8351E",
  "amino_acid" = "#852F88"
)

carbon_class_lookup <- bind_rows(auc_per_well, k_per_well) |>
  distinct(carbon_source, class) |>
  deframe()

label_colors <- unname(class_colors[carbon_class_lookup[carbon_levels]])
```

Each panel is one species at one streptomycin concentration. Species are never
compared across panels at matched dose, because no such dose exists.

```{r}
panel_levels <- c(
  "HAMBI_1287\n0 μg/ml Str",
  "HAMBI_1287\n512 μg/ml Str",
  "HAMBI_1977\n0 μg/ml Str",
  "HAMBI_1977\n10 μg/ml Str"
)

add_panel <- function(df) {
  df |>
    mutate(
      panel = factor(
        paste0(
          str_replace(strainID, "^H", "HAMBI_"),
          "\n",
          streptomycin_ug_ml,
          " μg/ml Str"
        ),
        levels = panel_levels
      )
    )
}
```

```{r}
plotter <- function(df, lab) {
  dw <- 0.5
  psize <- 3
  lsize <- 1

  df |>
    add_panel() |>
    mutate(carbon_source = factor(carbon_source, levels = carbon_levels)) |>
    # fill in the untested species/streptomycin/carbon combinations as explicit
    # NA rows. Without this the connecting line is drawn straight across the
    # four carbon sources that were never run with streptomycin, which reads as
    # an interpolated value rather than a gap.
    tidyr::complete(panel, evo_hist, carbon_source) |>
    ggplot(aes(y = mn, x = carbon_source, color = evo_hist)) +
    geom_linerange(
      aes(ymin = ci_lo, ymax = ci_hi),
      position = position_dodge(width = dw),
      size = lsize
    ) +
    geom_point(position = position_dodge(width = dw), size = psize) +
    geom_line(
      aes(group = evo_hist),
      alpha = 0.5,
      size = lsize - 0.5,
      position = position_dodge(width = dw),
    ) +
    # fixed rather than free_x: the +Str panels are missing four carbon
    # sources, and free scales would misalign the rows across panels
    facet_grid(~panel) +
    labs(
      y = lab,
      x = NULL,
      color = "Str Resistance",
    ) +
    scale_x_discrete(
      labels = function(x) {
        glue::glue("{{.{label_colors} {x}}}")
      }
    ) +
    scale_color_manual(values = c("#333333", "#EB990C")) +
    coord_flip() +
    theme_classic() +
    theme(
      axis.text.y = marquee::element_marquee(size = 11),
      strip.text = element_text(size = 12),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      strip.placement = "outside",
      strip.background = element_blank(),
      # upstream this legend sat inside the panel at (0.95, 0.95). With four
      # panels instead of two it lands on top of the right-hand panel's points,
      # so it moves out of the plotting area here.
      legend.position = "bottom",
      legend.key.size = unit(7, "mm"),
      legend.background = element_rect(
        fill = "grey95",
        colour = "grey20",
        linewidth = 0.5
      ),
      legend.title = element_blank()
    )
}
```

## Plot differences in mean maximum density (empirical K)

```{r}
pk <- plotter(k_summary, "Mean K (OD600)")

ggsave(
  here::here(figs, "one_carbon_K_strep.svg"),
  plot = pk,
  device = "svg",
  width = 13,
  height = 8
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 13
#| fig-height: 8
#| fig-cap: "Mean maximum density for ANC and EVO forms of each species grown on 20 substrates as the sole carbon source in a minimal medium lacking carbon, at each streptomycin concentration that species was dosed at. Mean K and bootstrapped 95% CI (2000 resamples) are computed for each combination of strainID, evo_hist, streptomycin_ug_ml and carbon_source. Point colour is streptomycin resistance evolution; y axis text colour is molecular class, starting at the top with sugars, then organic acids, then amino acids. Gaps in the +Str panels are the four carbon sources that were only run without streptomycin."
pk
```

## Plot differences in mean AUC (area under growth curve)

```{r}
pauc <- plotter(auc_summary, "Mean AUC (OD600•h)")

ggsave(
  here::here(figs, "one_carbon_AUC_strep.svg"),
  plot = pauc,
  device = "svg",
  width = 13,
  height = 8
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 13
#| fig-height: 8
#| fig-cap: "As in @fig-01 but for mean area under the growth curve."
pauc
```

## Plot differences in mean growth rate

```{r}
pgr <- plotter(gr_summary, "Mean μ (per hour)")

ggsave(
  here::here(figs, "one_carbon_mumax_strep.svg"),
  plot = pgr,
  device = "svg",
  width = 13,
  height = 8
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 13
#| fig-height: 8
#| fig-cap: "As in @fig-01 but for mean maximum per capita growth rate from the spline fits."
pgr
```
