---
title: "03.A Formatting co-culture growth curves from Logphase 600"
subtitle: "Quality control"
author: "Shane Hogle"
date: today
---

# Introduction

The ancestral (streptomycin sensitive) and evolved (streptomycin resistant) forms of HAMBI_1287 and HAMBI_1977 were grown on adjacent 0.22 μm membrane-separated wells in a Verasys co-culture plate. The membrane allows free diffusion of resources and metabolic byproducts but not cells. Note that due to the unique format of the plate rows A and H do not contain any data. The plates differ by streptomycin concentration as indicated below. Measurements were collected with a Biotek Logphase 600 microplate reader.

In this notebook we read the plate reader exports, join them to the samplesheets, smooth them, and inspect every well.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(fs)
library(ggforce)
library(slider)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw")
data_processed <- here::here("data", "processed", "03_coculture_plates")

# make processed data directory if it doesn't exist
fs::dir_create(data_processed)
```

# Batch01 - 2025.09.20

## Read sample metadata

```{r}
batch <- here::here(data_raw, "20250920_coculture_batch01")

samplesheet_diff_test <- readxl::read_xlsx(here::here(
  batch,
  "diffusion_test_2_samplesheet.xlsx"
))

samplesheet_coculture <- readxl::read_xlsx(here::here(
  batch,
  "coculture_samplesheet.xlsx"
))
```

## Read growth curves

Diffusion test plate01.

```{r}
plate01 <- read_logphase_xlsx(
  here::here(
    batch,
    "diffusion_test_2_Co-culture_1977_1287AE_12.9_15-syys-2025 09-00-26.xlsx"
  ),
  sheet = 2,
  skip = 1
) %>%
  # remove rows A and H
  filter(str_detect(well, "^A|^H", negate = TRUE)) %>%
  left_join(samplesheet_diff_test, by = join_by(well)) %>%
  mutate(plate_name = "plate01")
```

Experiment plate02 - 0 μg/ml streptomycin.

```{r}
plate02 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_0_4_8ug_plates_22-syys-2025 08-17-10.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  # rows A and H not collected here so they are NA
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 0) %>%
  mutate(plate_name = "plate02")
```

Experiment plate03 - 4 μg/ml streptomycin.

```{r}
plate03 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_0_4_8ug_plates_22-syys-2025 08-17-10.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 4) %>%
  mutate(plate_name = "plate03")
```

Experiment plate04 - 8 μg/ml streptomycin.

```{r}
plate04 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_0_4_8ug_plates_22-syys-2025 08-17-10.xlsx"),
  sheet = 8,
  skip = 1
) %>%
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 8) %>%
  mutate(plate_name = "plate04")
```

# Batch02 - 2025.09.25

## Read sample metadata

```{r}
batch <- here::here(data_raw, "20250925_coculture_batch02")

samplesheet_coculture <- readxl::read_xlsx(here::here(
  batch,
  "coculture_samplesheet.xlsx"
))
```

## Read growth curves

Experiment plate05 - 12 μg/ml streptomycin.

```{r}
plate05 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_12_16_24_ug_plates_25-syys-2025 13-42-42.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 12) %>%
  mutate(plate_name = "plate05")
```

Experiment plate06 - 16 μg/ml streptomycin.

```{r}
plate06 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_12_16_24_ug_plates_25-syys-2025 13-42-42.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 16) %>%
  mutate(plate_name = "plate06")
```

Experiment plate07 - 24 μg/ml streptomycin.

```{r}
plate07 <- read_logphase_xlsx(
  here::here(batch, "Co-culture_12_16_24_ug_plates_25-syys-2025 13-42-42.xlsx"),
  sheet = 8,
  skip = 1
) %>%
  drop_na() %>%
  left_join(samplesheet_coculture, by = join_by(well)) %>%
  mutate(streptomycin = 24) %>%
  mutate(plate_name = "plate07")
```

# Batch03 - 2025.10.06

Each plate in this batch has its own samplesheet.

```{r}
batch <- here::here(data_raw, "20251006_coculture_batch03")

plate08 <- read_logphase_xlsx(
  here::here(batch, "coculture_plates_8_9_10_11_06-10-2025.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  drop_na() %>%
  left_join(
    readxl::read_xlsx(here::here(batch, "coculture_samplesheet_plate08.xlsx")),
    by = join_by(well)
  ) %>%
  mutate(plate_name = "plate08")

plate09 <- read_logphase_xlsx(
  here::here(batch, "coculture_plates_8_9_10_11_06-10-2025.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  drop_na() %>%
  left_join(
    readxl::read_xlsx(here::here(batch, "coculture_samplesheet_plate09.xlsx")),
    by = join_by(well)
  ) %>%
  mutate(plate_name = "plate09")

plate10 <- read_logphase_xlsx(
  here::here(batch, "coculture_plates_8_9_10_11_06-10-2025.xlsx"),
  sheet = 8,
  skip = 1
) %>%
  drop_na() %>%
  left_join(
    readxl::read_xlsx(here::here(batch, "coculture_samplesheet_plate10.xlsx")),
    by = join_by(well)
  ) %>%
  mutate(plate_name = "plate10")

plate11 <- read_logphase_xlsx(
  here::here(batch, "coculture_plates_8_9_10_11_06-10-2025.xlsx"),
  sheet = 11,
  skip = 1
) %>%
  drop_na() %>%
  left_join(
    readxl::read_xlsx(here::here(batch, "coculture_samplesheet_plate11.xlsx")),
    by = join_by(well)
  ) %>%
  mutate(plate_name = "plate11")
```

# Batch04 - 2025.10.30

```{r}
batch <- here::here(data_raw, "20251030_coculture_batch04")

plate12 <- read_logphase_xlsx(
  here::here(batch, "coculture_plate12.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  drop_na() %>%
  left_join(
    readxl::read_xlsx(here::here(batch, "coculture_samplesheet_plate12.xlsx")),
    by = join_by(well)
  ) %>%
  mutate(plate_name = "plate12")
```

# Combine, tidy, format

Combine all samples, group by plate + well, calculate rolling mean. This reduces
the jaggedness of the curves and also makes the resulting saved data file
smaller.

```{r}
coculture_gcurves_sm <- bind_rows(
  plate01,
  plate02,
  plate03,
  plate04,
  plate05,
  plate06,
  plate07,
  plate08,
  plate09,
  plate10,
  plate11,
  plate12
) %>%
  dplyr::group_by(plate_name, well) %>%
  arrange(plate_name, well) %>%
  dplyr::mutate(
    OD600_rollmean = slider::slide_dbl(OD600, mean, .before = 2, .after = 2)
  ) %>%
  ungroup() %>%
  relocate(OD600_rollmean, .after = "OD600")

# save result for later
readr::write_tsv(
  coculture_gcurves_sm,
  here::here(data_processed, "gcurves_smooth.tsv")
)
```

# Inspect growth curves

```{r}
plotter <- function(id) {
  plotplate(
    coculture_gcurves_sm,
    dfxy = NULL,
    unsmoothed = TRUE,
    predicted = FALSE,
    plate = id,
    rows = 6,
    cols = 12,
    page = 1,
    scales = "fixed"
  )
}
```

## plate01

Growth curves for the "diffusion test" co-culture plate. Columns 1, 3, 5, and 7 have only M9 salts. Columns 2, 4, 6, 8, 9, 10, 11, and 12 had R2A medium. Bacteria were inoculated into rows B-D, columns 1, 3, 5, and 7 and rows E-F columns 2, 4, 6, and 8. Columns 9 and 11 were not inoculated with any bacteria.

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 12
#| fig-height: 8
#| fig-cap: Plate01 co-culture plate growth curves. Column $n=\{1\dots11\}$ is connected to column $n+1$ with a 0.22 μm membrane allowing free diffusion of resources and metabolic byproducts but not cells. X-axis is time in hours (48 hour incubation). Y axis is the absorbance scaled for each well. Blue line is smoothed with a moving average window of five points. Orange is non-smoothed raw data.
plotter("plate01")
```

## plate02

Growth curves from bacterial competition at 0 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate02 with 0 μg/ml streptomycin.
plotter("plate02")
```

## plate03

Growth curves from bacterial competition at 4 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate03 with 4 μg/ml streptomycin.
plotter("plate03")
```

## plate04

Growth curves from bacterial competition at 8 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-04
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate04 with 8 μg/ml streptomycin.
plotter("plate04")
```

## plate05

Growth curves from bacterial competition at 12 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-05
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate05 with 12 μg/ml streptomycin.
plotter("plate05")
```

## plate06

Growth curves from bacterial competition at 16 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-06
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate06 with 16 μg/ml streptomycin.
plotter("plate06")
```

## plate07

Growth curves from bacterial competition at 24 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-07
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate07 with 24 μg/ml streptomycin.
plotter("plate07")
```

## plate08

Growth curves from bacterial monocultures at 0, 4, 8 μg/ml streptomycin. Here each well pair contains only one bacterial strain.

```{r}
#| echo: false
#| output: true
#| label: fig-08
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate08.
plotter("plate08")
```

## plate09

Growth curves from bacterial monocultures at 12, 16, 24 μg/ml streptomycin. Here each well pair contains only one bacterial strain.

```{r}
#| echo: false
#| output: true
#| label: fig-09
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate09.
plotter("plate09")
```

## plate10

Growth curves from bacterial competition co-cultures at 64 and 256 μg/ml streptomycin.

```{r}
#| echo: false
#| output: true
#| label: fig-10
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate10.
plotter("plate10")
```

## plate11

- monocultures at 64 and 256 μg/ml streptomycin
- cocultures at 1028 and 4096 μg/ml streptomycin

```{r}
#| echo: false
#| output: true
#| label: fig-11
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate11.
plotter("plate11")
```

## plate12

- monocultures at 64 and 256 μg/ml streptomycin
- cocultures at 8 and 10 μg/ml streptomycin

```{r}
#| echo: false
#| output: true
#| label: fig-12
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-01 but for plate12.
plotter("plate12")
```
