---
title: "03.C Growth in pairs and monoculture"
subtitle: "Competition outcomes"
author: "Shane Hogle"
date: today
---

# Introduction

Strain naming convention: 1287 = *Citrobacter koseri* 1287, 1977 = *Pseudomonas chlororaphis* 1977, A = Ancestral/Streptomycin sensitive, E = Evolved/Streptomycin resistant.

Here we compare the growth of each strain in monoculture against its growth when sharing a membrane-separated well with a competitor, across a range of streptomycin concentrations.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(fs)
library(scales)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_processed <- here::here("data", "processed", "03_coculture_plates")
figs <- here::here("output", "figures")

# make output directories if they don't exist
fs::dir_create(data_processed)
fs::dir_create(figs)
```

## Functions

This function collects all the data from monocultures and cocultures for the
selected species pair. Only monoculture data from streptomycin concentrations
tested in co-cultures is included.

```{r}
collect_sp_pairs <- function(df, competition_pair, strep_thresh) {
  strainpairs <- stringr::str_split(competition_pair, "_")[[1]]
  df %>%
    mutate(streptomycin = if_else(plate_name == "plate01", 0, streptomycin)) %>%
    mutate(strainID = make_strainID(evolution, strain)) %>%
    filter(strainID %in% strainpairs) %>%
    filter(str_detect(culture_type, "diffusion", negate = T)) %>%
    filter(streptomycin <= {{ strep_thresh }}) %>%
    filter(
      (culture_type == "coculture" &
        competition_pair == {{ competition_pair }}) |
        culture_type == "monoculture"
    ) %>%
    mutate(
      culture_type = factor(
        culture_type,
        levels = c("monoculture", "coculture")
      )
    )
}
```

This function plots the output from the `collect_sp_pairs` function.

```{r}
plot_sp_pairs <- function(df, title) {
  df_cl <- df %>%
    mutate(seconds = seconds - seconds %% 1200) %>%
    mutate(hours = seconds / 3600) %>%
    summarize(
      y = mean(OD600_rollmean),
      sd = sd(OD600_rollmean),
      n = n(),
      .by = c(hours, strainID, streptomycin, competition_pair, culture_type)
    ) %>%
    mutate(ymin = y - 1.96 * sd / sqrt(n), ymax = y + 1.96 * sd / sqrt(n))

  ggplot2::ggplot() +
    ggplot2::geom_line(
      data = df_cl,
      aes(
        x = hours,
        y = y,
        color = strainID
      ),
    ) +
    ggplot2::geom_ribbon(
      data = df_cl,
      aes(
        x = hours,
        ymin = ymin,
        ymax = ymax,
        fill = strainID,
        group = interaction(strainID, streptomycin, culture_type)
      ),
      alpha = 0.5
    ) +
    ggplot2::geom_line(
      data = df,
      aes(
        x = hours,
        y = OD600_rollmean,
        color = strainID,
        group = interaction(well, plate_name)
      ),
      alpha = 0.25,
      linewidth = 0.25
    ) +
    ggplot2::labs(x = "Hours", y = "OD600") +
    ggplot2::scale_x_continuous(
      breaks = seq(0, 48, 12),
      labels = seq(0, 48, 12)
    ) +
    ggplot2::labs(
      y = "Optical density (600 nm), log scale",
      x = "Hours",
      title = {{ title }}
    ) +
    ggplot2::scale_color_manual(values = straincols) +
    ggplot2::scale_fill_manual(values = straincols) +
    ggplot2::scale_y_continuous(trans = "log", breaks = c(0, 0.25, 0.5, 1)) +
    ggplot2::facet_grid(culture_type ~ streptomycin) +
    ggplot2::theme_bw() +
    theme(legend.position = "bottom")
}
```

## Read data

```{r}
gcurves <- readr::read_tsv(here::here(data_processed, "gcurves_smooth.tsv"))
```

# Diffusion test

Plot the co-cultures growing on M9 compared to the control (column 1 and 3). On the plate, column $n=\{1\dots11\}$ is connected to column $n+1$ with a 0.22 μm membrane allowing free diffusion of resources and metabolic byproducts but not cells. These experiments were all conducted with no streptomycin.

```{r}
pdiffusiontest <- gcurves %>%
  filter(str_detect(culture_type, "diffusion")) %>%
  filter(str_detect(strain, "NANA", negate = TRUE)) %>%
  ggplot() +
  ggplot2::geom_line(aes(
    x = hours,
    y = OD600_rollmean,
    lty = replicate,
    color = culture_type,
    group = interaction(well, plate_name)
  )) +
  ggplot2::labs(x = "Hours", y = "OD600") +
  ggplot2::scale_x_continuous(
    breaks = seq(0, 48, 12),
    labels = seq(0, 48, 12)
  ) +
  labs(y = "Optical density (600 nm)", x = "Hours", title = "Diffusion test") +
  facet_grid(evolution ~ strain) +
  theme(legend.position = "bottom")

ggsave(
  here::here(figs, "coculture_diffusion_test.svg"),
  plot = pdiffusiontest,
  device = "svg",
  width = 7,
  height = 6
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 7
#| fig-height: 6
#| fig-cap: "Growth curves (OD600 over 48 hours) in monoculture for three replicates (line type) of each species (plot panel columns) and evolution status (plot panel rows). Line color indicates experimental conditions: inoculation into M9 salts with an adjacent well containing sterile R2A growth medium, versus the inverse where cells are inoculated into sterile R2A with an adjacent well containing sterile M9 salts."
pdiffusiontest
```

# Co-culture and monoculture growth

## A1287 + E1287

```{r}
A1287E1287 <- collect_sp_pairs(
  gcurves,
  competition_pair = "A1287_E1287",
  strep_thresh = 256
) %>%
  filter(streptomycin != 10) %>%
  write_tsv(here::here(data_processed, "competition_A1287_E1287_final.tsv"))

pA1287E1287 <- plot_sp_pairs(
  A1287E1287,
  title = "Sensitive 1287 (A1287) + Resistant 1287 (E1287)"
)

ggsave(
  here::here(figs, "A1287_E1287.svg"),
  plot = pA1287E1287,
  device = "svg",
  width = 9,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 9
#| fig-height: 4
#| fig-cap: Growth curves (OD600 over 48 hours) of two strains in a competing pair (denoted by color and plot title). Columns in the plot grid depict different streptomycin concentrations (μg/ml) under which growth was measured. Grid rows depict whether strains were grown in co-culture (strains inoculated into adjacent wells) or in monoculture (single strain inoculated into well with adjacent well containing sterile media only). Heavy line corresponds to mean growth while lighter lines correspond to individual replicates. The ribbon around each mean line corresponds to the 95% confidence interval (CI) around the mean. Line color corresponds to species/strain identity.
pA1287E1287
```

## A1977 + E1977

```{r}
A1977E1977 <- collect_sp_pairs(
  gcurves,
  competition_pair = "A1977_E1977",
  strep_thresh = 24
) %>%
  # no co-cultures for these pairs done at 10 ug/ml
  filter(streptomycin != 10) %>%
  # filter out the strange streptomycin 8 ug/ml monoculture samples
  filter(
    !(streptomycin == 8 & plate_name == "plate08" & strainID == "A1977")
  ) %>%
  write_tsv(here::here(data_processed, "competition_A1977_E1977_final.tsv"))

pA1977E1977 <- plot_sp_pairs(
  A1977E1977,
  title = "Sensitive 1977 (A1977) + Resistant 1977 (E1977)"
)

ggsave(
  here::here(figs, "A1977_E1977.svg"),
  plot = pA1977E1977,
  device = "svg",
  width = 8,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 8
#| fig-height: 4
#| fig-cap: As in @fig-02 but for A1977 and E1977.
pA1977E1977
```

## A1287 + A1977

```{r}
A1287A1977 <- collect_sp_pairs(
  gcurves,
  competition_pair = "A1287_A1977",
  strep_thresh = 24
) %>%
  # filter out the strange streptomycin 8 ug/ml monoculture samples
  filter(
    !(streptomycin == 8 & plate_name == "plate08" & strainID == "A1977")
  ) %>%
  write_tsv(here::here(data_processed, "competition_A1287_A1977_final.tsv"))

pA1287A1977 <- plot_sp_pairs(
  A1287A1977,
  title = "Sensitive 1287 (A1287) + Sensitive 1977 (A1977)"
)

ggsave(
  here::here(figs, "A1287_A1977.svg"),
  plot = pA1287A1977,
  device = "svg",
  width = 8,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-04
#| fig-width: 8
#| fig-height: 4
#| fig-cap: As in @fig-02 but for A1287 and A1977.
pA1287A1977
```

## E1287 + A1977

```{r}
E1287A1977 <- collect_sp_pairs(
  gcurves,
  competition_pair = "E1287_A1977",
  strep_thresh = 24
) %>%
  # no co-cultures for these pairs done at 10 ug/ml
  filter(streptomycin != 10) %>%
  # filter out the strange streptomycin 8 ug/ml monoculture samples
  filter(
    !(streptomycin == 8 & plate_name == "plate08" & strainID == "A1977")
  ) %>%
  write_tsv(here::here(data_processed, "competition_E1287_A1977_final.tsv"))

pE1287A1977 <- plot_sp_pairs(
  E1287A1977,
  title = "Resistant 1287 (E1287) + Sensitive 1977 (A1977)"
)

ggsave(
  here::here(figs, "E1287_A1977.svg"),
  plot = pE1287A1977,
  device = "svg",
  width = 8,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-05
#| fig-width: 8
#| fig-height: 4
#| fig-cap: As in @fig-02 but for E1287 and A1977.
pE1287A1977
```

## A1287 + E1977

```{r}
A1287E1977 <- collect_sp_pairs(
  gcurves,
  competition_pair = "A1287_E1977",
  strep_thresh = 256
) %>%
  # no co-cultures for these pairs done at 10 ug/ml
  filter(streptomycin != 10) %>%
  # filter out the strange streptomycin 8 ug/ml monoculture samples
  filter(
    !(streptomycin == 8 & plate_name == "plate08" & strainID == "A1977")
  ) %>%
  write_tsv(here::here(data_processed, "competition_A1287_E1977_final.tsv"))

pA1287E1977 <- plot_sp_pairs(
  A1287E1977,
  title = "Sensitive 1287 (A1287) + Resistant 1977 (E1977)"
)

ggsave(
  here::here(figs, "A1287_E1977.svg"),
  plot = pA1287E1977,
  device = "svg",
  width = 9,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-06
#| fig-width: 9
#| fig-height: 4
#| fig-cap: As in @fig-02 but for A1287 and E1977.
pA1287E1977
```

## E1287 + E1977

```{r}
E1287E1977 <- collect_sp_pairs(
  gcurves,
  competition_pair = "E1287_E1977",
  strep_thresh = 8025600
) %>%
  filter(streptomycin != 10) %>%
  # filter out the strange streptomycin 8 ug/ml monoculture samples
  filter(
    !(streptomycin == 8 & plate_name == "plate08" & strainID == "A1977")
  ) %>%
  # removes a bad replicate where OD is over 0.8 at time 0
  filter(
    !(culture_type == "monoculture" &
      plate_name == "plate11" &
      joined_well_pair == "E07E08")
  ) %>%
  write_tsv(here::here(data_processed, "competition_E1287_E1977_final.tsv"))

pE1287E1977 <- plot_sp_pairs(
  E1287E1977,
  title = "Resistant 1287 (E1287) + Resistant 1977 (E1977)"
)

ggsave(
  here::here(figs, "E1287_E1977.svg"),
  plot = pE1287E1977,
  device = "svg",
  width = 10,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-07
#| fig-width: 10
#| fig-height: 4
#| fig-cap: As in @fig-02 but for E1287 and E1977.
pE1287E1977
```
