---
title: "04.B Formatting serial transfer growth curves"
subtitle: "Quality control, 2026-04-10 experiment"
author: "Shane Hogle"
date: today
---

# Introduction

Strain naming convention: 1287 = *Citrobacter koseri* 1287, 1977 = *Pseudomonas chlororaphis* 1977, A = Ancestral/Streptomycin sensitive, E = Evolved/Streptomycin resistant.

Growth media convention: M9MM = M9 minimal media with glucose as the sole carbon source and supplemented with casamino acids. R2A = Reasoner's 2 broth (complex growth medium).

This experiment ran for five 48-hour growth cycles (240 hours of growth total) where each well was serially diluted to a concentration of 10^-4^ (10,000 fold dilution) and then placed back in the incubating plate reader.

In this notebook we read the five cycle exports, join them to the samplesheet, and smooth them. Each cycle export holds the 1977 plate on sheet 2 and the 1287 plate on sheet 5.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
library(fs)
library(slider)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw", "20260410_serial_transfer")
data_processed <- here::here(
  "data",
  "processed",
  "04_serial_transfer_coculture"
)

# make processed data directory if it doesn't exist
fs::dir_create(data_processed)
```

## Read sample metadata

```{r}
samplesheet_coculture <- readxl::read_xlsx(here::here(
  data_raw,
  "coculture_samplesheet.xlsx"
))
```

All dilutions done at 10,000 fold.

# Cycle 1

```{r}
c1_1977 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_1_start_10-huhti-2026 12-14-59.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1977"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c1", start_hrs_from_0 = 0)

c1_1287 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_1_start_10-huhti-2026 12-14-59.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1287"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c1", start_hrs_from_0 = 0)
```

# Cycle 2

```{r}
c2_1977 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_2_10.4_12-huhti-2026 13-12-32.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1977"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c2", start_hrs_from_0 = 48)

c2_1287 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_2_10.4_12-huhti-2026 13-12-32.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1287"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c2", start_hrs_from_0 = 48)
```

# Cycle 3

```{r}
c3_1977 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_3_12.4_14-huhti-2026 13-05-36.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1977"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c3", start_hrs_from_0 = 96)

c3_1287 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_3_12.4_14-huhti-2026 13-05-36.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1287"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c3", start_hrs_from_0 = 96)
```

# Cycle 4

```{r}
c4_1977 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_4_16-huhti-2026 13-33-18.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1977"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c4", start_hrs_from_0 = 144)

c4_1287 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_4_16-huhti-2026 13-33-18.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1287"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c4", start_hrs_from_0 = 144)
```

# Cycle 5

```{r}
c5_1977 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_5_20-huhti-2026 07-42-52.xlsx"),
  sheet = 2,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1977"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c5", start_hrs_from_0 = 192)

c5_1287 <- read_logphase_xlsx(
  here::here(data_raw, "Serial_transfer_5_20-huhti-2026 07-42-52.xlsx"),
  sheet = 5,
  skip = 1
) %>%
  right_join(
    filter(samplesheet_coculture, strain == "H1287"),
    by = join_by(well)
  ) %>%
  mutate(dilution = 10000, plate_name = "c5", start_hrs_from_0 = 192)
```

# Combine, tidy, format

Combine all samples, group by plate + well, calculate rolling mean. This reduces
the jaggedness of the curves and also makes the resulting saved data file
smaller.

```{r}
smooth_cycles <- function(...) {
  bind_rows(...) %>%
    dplyr::group_by(plate_name, well) %>%
    arrange(plate_name, well) %>%
    dplyr::mutate(
      OD600_rollmean = slider::slide_dbl(OD600, mean, .before = 2, .after = 2)
    ) %>%
    ungroup() %>%
    relocate(OD600_rollmean, .after = "OD600")
}

coculture_gcurves_sm_1977 <- smooth_cycles(
  c1_1977,
  c2_1977,
  c3_1977,
  c4_1977,
  c5_1977
)

coculture_gcurves_sm_1287 <- smooth_cycles(
  c1_1287,
  c2_1287,
  c3_1287,
  c4_1287,
  c5_1287
)
```

```{r}
readr::write_tsv(
  coculture_gcurves_sm_1977,
  here::here(data_processed, "20260410_1977_gcurves_smooth.tsv")
)

readr::write_tsv(
  coculture_gcurves_sm_1287,
  here::here(data_processed, "20260410_1287_gcurves_smooth.tsv")
)
```

The per-cycle plate grids and the competition time courses are plotted in the
[next notebook](03_20260410_prune_and_plot.qmd).
