---
title: "04.C Serial transfer of co-culture pairs"
subtitle: "Pruning and competition time courses, 2026-04-10 experiment"
author: "Shane Hogle"
date: today
---

# Introduction

Strain naming convention: 1287 = *Citrobacter koseri* 1287, 1977 = *Pseudomonas chlororaphis* 1977, A = Ancestral/Streptomycin sensitive, E = Evolved/Streptomycin resistant.

Growth media convention: M9MM = M9 minimal media with glucose as the sole carbon source and supplemented with casamino acids. R2A = Reasoner's 2 broth (complex growth medium).

This experiment ran for five 48-hour growth cycles (240 hours of growth total) where each well was serially diluted to a concentration of 10^-4^ (10,000 fold dilution) and then placed back in the incubating plate reader.

Here we inspect every cycle, drop the well pairs that grew where nothing should have grown, and plot the competition time course with all five cycles joined end to end.

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(fs)
library(ggforce)
source(here::here("R", "gcurves.R"))
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_processed <- here::here(
  "data",
  "processed",
  "04_serial_transfer_coculture"
)
figs <- here::here("output", "figures")

# make output directories if they don't exist
fs::dir_create(data_processed)
fs::dir_create(figs)
```

## Read data

Formatted in the [prior step](02_20260410_format_growthcurves.qmd).

```{r}
coculture_gcurves_sm_1977 <- readr::read_tsv(
  here::here(data_processed, "20260410_1977_gcurves_smooth.tsv")
)

coculture_gcurves_sm_1287 <- readr::read_tsv(
  here::here(data_processed, "20260410_1287_gcurves_smooth.tsv")
)
```

# Inspect individual growth curves

```{r}
plotter <- function(id, df) {
  plotplate(
    df,
    dfxy = NULL,
    unsmoothed = TRUE,
    predicted = FALSE,
    plate = id,
    rows = 6,
    cols = 12,
    page = 1,
    scales = "fixed"
  )
}
```

## *Pseudomonas chlororaphis* 1977

::: {.panel-tabset}

### Cycle 1

```{r}
#| echo: false
#| output: true
#| label: fig-1977-c1
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Growth curves for *P. chlororaphis* 1977, cycle 1. X-axis is time in hours (48 hour cycle). Blue line is smoothed with a moving average window of five points, orange dashed is non-smoothed raw data."
plotter("c1", coculture_gcurves_sm_1977)
```

### Cycle 2

```{r}
#| echo: false
#| output: true
#| label: fig-1977-c2
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1977-c1 but for cycle 2.
plotter("c2", coculture_gcurves_sm_1977)
```

### Cycle 3

```{r}
#| echo: false
#| output: true
#| label: fig-1977-c3
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1977-c1 but for cycle 3.
plotter("c3", coculture_gcurves_sm_1977)
```

### Cycle 4

```{r}
#| echo: false
#| output: true
#| label: fig-1977-c4
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1977-c1 but for cycle 4.
plotter("c4", coculture_gcurves_sm_1977)
```

### Cycle 5

```{r}
#| echo: false
#| output: true
#| label: fig-1977-c5
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1977-c1 but for cycle 5.
plotter("c5", coculture_gcurves_sm_1977)
```

:::

## *Citrobacter koseri* 1287

::: {.panel-tabset}

### Cycle 1

```{r}
#| echo: false
#| output: true
#| label: fig-1287-c1
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Growth curves for *C. koseri* 1287, cycle 1. X-axis is time in hours (48 hour cycle). Blue line is smoothed with a moving average window of five points, orange dashed is non-smoothed raw data."
plotter("c1", coculture_gcurves_sm_1287)
```

### Cycle 2

```{r}
#| echo: false
#| output: true
#| label: fig-1287-c2
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1287-c1 but for cycle 2.
plotter("c2", coculture_gcurves_sm_1287)
```

### Cycle 3

```{r}
#| echo: false
#| output: true
#| label: fig-1287-c3
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1287-c1 but for cycle 3.
plotter("c3", coculture_gcurves_sm_1287)
```

### Cycle 4

```{r}
#| echo: false
#| output: true
#| label: fig-1287-c4
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1287-c1 but for cycle 4.
plotter("c4", coculture_gcurves_sm_1287)
```

### Cycle 5

```{r}
#| echo: false
#| output: true
#| label: fig-1287-c5
#| fig-width: 12
#| fig-height: 8
#| fig-cap: As in @fig-1287-c1 but for cycle 5.
plotter("c5", coculture_gcurves_sm_1287)
```

:::

# Blacklists

These are well pairs where there was growth where there should be none,
identified by inspecting the plots above.

```{r}
blacklist_1287 <- tribble(
  ~strainID , ~plate_name , ~joined_well_pair ,
  "H1287"   , "c2"        , "C05C06"          ,
  "H1287"   , "c2"        , "E05E06"          ,
  "H1287"   , "c2"        , "F05F06"          ,
  "H1287"   , "c2"        , "E09E10"          ,
  "H1287"   , "c3"        , "C05C06"          ,
  "H1287"   , "c3"        , "E09E10"          ,
  "H1287"   , "c4"        , "C05C06"          ,
  "H1287"   , "c4"        , "E09E10"          ,
  "H1287"   , "c5"        , "C05C06"          ,
  "H1287"   , "c5"        , "E09E10"
)

blacklist_1977 <- tribble(
  ~strainID , ~plate_name , ~joined_well_pair ,
  "H1977"   , "c2"        , "D05D06"          ,
  "H1977"   , "c2"        , "E05E06"          ,
  "H1977"   , "c3"        , "D05D06"          ,
  "H1977"   , "c3"        , "E05E06"          ,
  "H1977"   , "c3"        , "G05G06"          ,
  "H1977"   , "c3"        , "B09B10"          ,
  "H1977"   , "c3"        , "C09C10"          ,
  "H1977"   , "c4"        , "D05D06"          ,
  "H1977"   , "c4"        , "E05E06"          ,
  "H1977"   , "c4"        , "G05G06"          ,
  "H1977"   , "c4"        , "B09B10"          ,
  "H1977"   , "c4"        , "C09C10"          ,
  "H1977"   , "c5"        , "D05D06"          ,
  "H1977"   , "c5"        , "E05E06"          ,
  "H1977"   , "c5"        , "G05G06"          ,
  "H1977"   , "c5"        , "B09B10"          ,
  "H1977"   , "c5"        , "C09C10"
)
```

::: {.callout-note}
Now that I think about this there is no need to include monocultures on these
plates, because we have that information from the previous non-serial transfer
runs. This might save us some plates in the future.
:::

Drop the blacklisted well pairs and keep only the wells that actually hold a
culture. The pruned tables are written under their own names so that re-running
this notebook never destroys the output of the formatting step.

```{r}
prune_wells <- function(df, blacklist) {
  df |>
    anti_join(blacklist, by = join_by(joined_well_pair, plate_name)) |>
    filter(
      (culture_type == "monoculture" &
        str_detect(well, "[BCD]01|[BCD]05|[BCD]09") |
        str_detect(well, "[EFG]02|[EFG]06|[EFG]10")) |
        culture_type == "coculture"
    )
}

coculture_gcurves_pruned_1287 <- prune_wells(
  coculture_gcurves_sm_1287,
  blacklist_1287
)

coculture_gcurves_pruned_1977 <- prune_wells(
  coculture_gcurves_sm_1977,
  blacklist_1977
)

readr::write_tsv(
  coculture_gcurves_pruned_1287,
  here::here(data_processed, "20260410_1287_gcurves_pruned.tsv")
)

readr::write_tsv(
  coculture_gcurves_pruned_1977,
  here::here(data_processed, "20260410_1977_gcurves_pruned.tsv")
)
```

# Growth curves with transfers combined

```{r}
cyclelines <- tibble(
  plate_name = c("c1", "c2", "c3", "c4", "c5"),
  cycle_hour = c(48, 96, 144, 192, 240)
)

plot_competition_timecourse <- function(df, title) {
  df |>
    mutate(hours_agg = hours + start_hrs_from_0) %>%
    mutate(
      strainID = paste0(
        str_to_upper(str_sub(evolution, 1, 1)),
        str_sub(strain, 2, 5)
      ),
      growthenv = paste(media, strep_conc_ugml, sep = "_")
    ) %>%
    mutate(
      strainID = factor(
        strainID,
        levels = c("A1287", "E1287", "A1977", "E1977")
      )
    ) %>%
    summarize(
      y = mean(OD600_rollmean),
      sd = sd(OD600_rollmean),
      .by = c(
        hours_agg,
        strainID,
        evolution,
        growthenv,
        dilution,
        culture_type
      )
    ) %>%
    ggplot(aes(
      x = hours_agg,
    )) +
    geom_vline(data = cyclelines, aes(xintercept = cycle_hour), lty = 3) +
    ggplot2::geom_ribbon(
      aes(ymin = y - sd, ymax = y + sd, fill = factor(evolution)),
      alpha = 0.5,
      show.legend = FALSE
    ) +
    ggplot2::geom_line(
      aes(y = y, color = factor(evolution)),
      linewidth = 0.75
    ) +
    facet_grid(growthenv ~ culture_type) +
    labs(
      y = "OD",
      x = "Total hours from start",
      color = "Evolution",
      title = {{ title }}
    )
}
```

## *Citrobacter koseri* 1287

```{r}
p1287 <- plot_competition_timecourse(
  coculture_gcurves_pruned_1287,
  title = "Citrobacter 1287"
)

ggsave(
  here::here(figs, "20260410_1287_competition_timecourse.svg"),
  p1287,
  device = "svg",
  dpi = 300,
  width = 10,
  height = 6
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 10
#| fig-height: 6
#| fig-cap: Serial transfer growth curves (OD600) over five 48 hour growth cycles (240 hours total) for *Citrobacter koseri* 1287. Columns in the plot grid depict growth in monoculture or co-culture. Grid rows depict the growth medium and streptomycin concentrations. Heavy line corresponds to mean growth and the ribbon around each mean line corresponds to one standard deviation. Dashed vertical lines depict the serial transfer events. Color depicts the streptomycin resistance status (evo = resistant, anc = streptomycin sensitive).
p1287
```

## *Pseudomonas chlororaphis* 1977

```{r}
p1977 <- plot_competition_timecourse(
  coculture_gcurves_pruned_1977,
  title = "Pseudomonas 1977"
)

ggsave(
  here::here(figs, "20260410_1977_competition_timecourse.svg"),
  p1977,
  device = "svg",
  dpi = 300,
  width = 10,
  height = 6
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 10
#| fig-height: 6
#| fig-cap: As in @fig-01 but for *Pseudomonas chlororaphis* 1977.
p1977
```
