---
title: "01.A Monoculture growth at a range of streptomycin concentrations"
subtitle: "Bioscreen growth curves"
author: "Shane Hogle"
date: today
---

# Introduction

Inspect the monoculture growth of both Streptomycin-sensitive (Sens/Anc) and Streptomycin-resistant (Res/Evo) forms of *Citrobacter koseri* 1287 and *Pseudomonas chlororaphis* 1977 at different streptomycin concentrations.

These are high resolution growth curves (every 10 minutes) collected on a very nice instrument (Bioscreen). Here we examine how the monoculture growth changed with respect to evolutionary history and streptomycin concentrations.

::: {.callout-note}
## Provenance of the input data

The three TSVs read below are not produced by any notebook in this repository. They are an export from the sister project [hambiEvoEnvCoexist](https://github.com/slhogle/hambiEvoEnvCoexist), subset to the two species used here, and are treated as immutable raw input. The formatting and growth-rate estimation that produced them lives in `scripts/01_r2a_gcurves_bioscreen/` of that project.
:::

# Setup

## Libraries

```{r}
library(here)
library(tidyverse)
library(fs)
library(scales)
source(here::here("R", "generic.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw", "20240328_bioscreen_strains")
data_processed <- here::here("data", "processed", "01_bioscreen_strains")
figs <- here::here("output", "figures")

# make processed data and figure directories if they don't exist
fs::dir_create(data_processed)
fs::dir_create(figs)
```

## Read data

```{r}
many_auc_res <- readr::read_tsv(here::here(data_raw, "gcurve_auc_results.tsv"))

many_spline_res <- readr::read_tsv(here::here(
  data_raw,
  "gcurve_spline_results.tsv"
))

gcurves <- readr::read_tsv(here::here(
  data_raw,
  "gcurves_formatted_thinned.tsv"
))
```

# Growth curves

```{r}
pgcurves <- gcurves %>%
  mutate(
    hist = str_split_i(sp_hist, "_", 1),
    sp = str_split_i(sp_hist, "_", 2),
    strep_conc = factor(strep_conc, ordered = TRUE)
  ) %>%
  mutate(sp = sp_labels[sp]) %>%
  ggplot(aes(x = hours, y = OD600_rollmean, color = hist)) +
  geom_line(aes(group = interaction(sp_hist, bioscreen_well))) +
  labs(
    y = "Optical density (OD600)",
    x = "Time (hours)",
    color = "Evolutionary history"
  ) +
  facet_grid(strep_conc ~ sp) +
  scale_color_manual(values = histcols) +
  theme_mct()

ggsave(
  here::here(figs, "growthcurves_monoculture.svg"),
  plot = pgcurves,
  device = "svg",
  width = 7,
  height = 12
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-01
#| fig-width: 7
#| fig-height: 12
#| fig-cap: Growth curves of *Citrobacter koseri* 1287 (left column grid) and *Pseudomonas chlororaphis* 1977 (right column grid) at different streptomycin concentrations (grid rows) and evolutionary histories/sensitivities to streptomycin (colors). Each line is a biological replicate (n=5). Growth (culture turbidity) is measured at 10 minute intervals over 48 hours.
pgcurves
```

# Growth rates

```{r}
gr_strep <- many_spline_res %>%
  mutate(
    hist = str_split_i(sp_hist, "_", 1),
    sp = str_split_i(sp_hist, "_", 2)
  ) %>%
  summarize(ggplot2::mean_cl_boot(mumax), .by = c(sp, hist, strep_conc)) %>%
  mutate(
    sp = sp_labels[sp],
    strep_conc = factor(strep_conc, ordered = TRUE)
  ) %>%
  ggplot(aes(x = strep_conc, y = y)) +
  geom_linerange(aes(ymin = ymin, ymax = ymax, color = hist)) +
  geom_line(aes(color = hist, group = hist), lty = 1) +
  geom_point(aes(color = hist)) +
  labs(
    y = "Maximum per capita growth rate μ (hr-1)",
    x = "Streptomycin conc. (μg/ml)",
    color = "Evolutionary\nhistory"
  ) +
  facet_grid(~sp) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_color_manual(values = histcols) +
  theme_mct()

ggsave(
  here::here(figs, "u_streptomycin_monoculture.svg"),
  plot = gr_strep,
  device = "svg",
  width = 7,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-02
#| fig-width: 7
#| fig-height: 4
#| fig-cap: Maximum specific growth rate ($\mathrm{\mu_{max}}$ in $\mathrm{hr^{-1}}$) for the two bacterial species (panels) and their different evolutionary histories (colors) at different streptomycin concentrations (horizontal axis, μg/ml). Points are the mean over five biological replicates and line ranges show the 95% confidence interval from bootstraps. evo = experimentally evolved to high streptomycin.
gr_strep
```

# Maximum carrying capacity (K, maximum observed OD)

```{r}
k_strep <- many_auc_res %>%
  mutate(
    hist = str_split_i(sp_hist, "_", 1),
    sp = str_split_i(sp_hist, "_", 2)
  ) %>%
  summarize(ggplot2::mean_cl_boot(max_od), .by = c(sp, hist, strep_conc)) %>%
  mutate(
    sp = sp_labels[sp],
    strep_conc = factor(strep_conc, ordered = TRUE)
  ) %>%
  ggplot(aes(x = strep_conc, y = y)) +
  geom_linerange(aes(ymin = ymin, ymax = ymax, color = hist)) +
  geom_line(aes(color = hist, group = hist), lty = 1) +
  geom_point(aes(color = hist)) +
  labs(
    y = "Maximum carrying capacity (K, OD units)",
    x = "Streptomycin conc. (μg/ml)",
    color = "Evolutionary\nhistory"
  ) +
  facet_grid(~sp) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_color_manual(values = histcols) +
  theme_mct()

ggsave(
  here::here(figs, "k_streptomycin_monoculture.svg"),
  plot = k_strep,
  device = "svg",
  width = 7,
  height = 4
)
```

```{r}
#| echo: false
#| output: true
#| label: fig-03
#| fig-width: 7
#| fig-height: 4
#| fig-cap: As in @fig-02 but for maximum carrying capacity (K, maximum observed OD).
k_strep
```

# Area under the curve (AUC) vs streptomycin level

Not run, but retained for reference.

```{r}
#| eval: false
auc_strep <- many_auc_res %>%
  mutate(
    hist = str_split_i(sp_hist, "_", 1),
    sp = str_split_i(sp_hist, "_", 2)
  ) %>%
  summarize(ggplot2::mean_cl_boot(auc), .by = c(sp, hist, strep_conc)) %>%
  mutate(
    sp = sp_labels[sp],
    strep_conc = factor(strep_conc, ordered = TRUE)
  ) %>%
  ggplot(aes(x = strep_conc, y = y)) +
  geom_linerange(aes(ymin = ymin, ymax = ymax, color = hist)) +
  geom_line(aes(color = hist, group = hist), lty = 1) +
  geom_point(aes(color = hist)) +
  labs(
    y = "Total area under the growth curve",
    x = "Streptomycin conc. (μg/ml)",
    color = "Evolutionary\nhistory"
  ) +
  facet_grid(~sp) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_color_manual(values = histcols) +
  theme_mct()
```
